---
title: "ClimateNA_data_pull"
author: "Julia Harenčár"
format: 
  html:
    code-fold: true
    code-tools: true
    code-overflow: wrap
    toc: true
    number-sections: true
theme:
  light: minty
  dark: superhero
editor_options: 
  chunk_output_type: console
---

## Pulling ClimateNA Data

This script is modified from XXX . It uses GPS coordinates and a calculated 'lived climate' time period for each specimen to pull data for each herbarium specimen independently.

First, set working directory and load packages.

```{r}
#| include: false
# load packages
library(tidyverse)
library(cowplot) # for plot aesthetics
library(magrittr)
library(ggplot2)
theme_set(theme_bw())
library(ggmap)
library(htmlwidgets)
library(leaflet)
library(leaflet.extras)
library(RColorBrewer)
```

### Generate ClimateNA input file

Then, we load the specimen data and make a new csv with coordinates and IDs in the format required by ClimateNA.

```{r}
#| label: climate input data
#| eval: false
#| include: false

# load the data
specimens <- read.csv("Input_Files/FullSpecimenCCH2Data.csv", header = T)

# make specific dataframe for pulling data from climateNA
# ClimateNA requires an input csv containing the columns ID1, ID2, lat, long, and (optionally) el.
# ID1 and ID2 can be anything you want 
# currently not including elevation, but could calculate this from the gps and add it in later

locs <- specimens %>% 
  mutate(id2 = NA,
         elev = NA) %>% # add a blank ID2 col
  dplyr::select(id1 = id, id2, lat = decimalLatitude, long = decimalLongitude, elev)

head(locs) # peak at the new df to check; looks good!

# Write out the new df as a csv for input into the ClimateNA program
write.csv(locs[,], "climateNA_locations.csv", row.names = FALSE, quote = FALSE)
# NOTE!! This csv does NOT work for pulling the location data because it is in Mac not windows format. I solved by copying all the data EXCEPT the last, empty line (using sublime) from this saved .csv location file into a copy of the test location file included in the ClimateNA download. Hypothetically I should be able to convert the format of my file, but my one attempt failed and the copy paste worked.
```

#### Get ClimateNA data

I downloaded the ClimateNA program and wine to run on mac. I run the below command line prompt to

```{bash}
# open ClimateNA gui on my mac
  # arch -x86_64 wine Z:\\Users\\juliaharencar\\Documents\\Software\\ClimateNA_v760\\ClimateNA_v7.60.exe
```

To get the data, I select "Historical time series" in the dropdown menu under "multiple locations". Select "monthly all variables (180)". Specify the input file we just created above and an output file. Generate monthly climate data for the time window of interest. Save output file. [Definitions of each variable acronym can be found here.](https://climatebc.ca/help/climateBC/help.htm#_Toc410137604)

Note, growing degree days marked as days \> 5°C: DD5_01 – DD5_12     January - December degree-days above 5°C

### Reformat ClimateNA data

NOTE - eval is false here because generating the all_data df takes a long time (like, minutes), so we will only run when we need to modify the saved output climateNA_full_data_tall.csv.

**Climate Variable Key:**

```{r}
#| label: climate variable key
#Create a Table from a CSV to serve as a Key to Climate variables
#install.packages("knitr")
library(knitr)
legend <- read.csv("Figures/Climate_variable_Legend.csv")
kable(legend, format = "markdown")
```

**Code for reformatting Climate Data for our needs:**

```{r}
#| label: reformat data
#| eval: false
#| include: false

# 3. Reformat ClimateNA data ----------------------------------------------

# Data is formatted with one row per site/year, one column per variable/month

all_data <- read_csv("Input_Files/Raph_locs_1901-2024Mfull.csv") %>%
  dplyr::select(-id2,-Elevation) %>%
  # Make data tall (one row per variable/site/month/year)
  gather(variable, value, -Year, -id1, -Latitude, -Longitude) %>%
  # Split month from variable name
  separate(variable, -2, into = c("variable", "month"), convert = TRUE) %>%
  # Spread again so that there is one row per site/year/month, one column per variable type
  spread(variable, value) %>%
  # Make a climate date column
  mutate(Year = as.numeric(Year),
         month = as.numeric(month),
         month_pad = str_pad(month, 2, pad = "0"),
         clim_date = as.numeric(paste0(Year, month_pad))) %>%
  # Drop unnecessary columns
  select(-month_pad) %>%
  # Rename to standard variable names
  rename(id = id1, longitude = Longitude, latitude = Latitude, clim_year = Year, clim_month = month, ppt_mm = PPT, srad = rsds, tmin = Tmin, tave = Tave, tmax = Tmax, CMD_mm = CMD, CMI_mm = CMI, chillDD = DD_0_, heatDD = DD_18_, coolDD = DD18_, growDD= DD5_, FFD = NFFD)

# Rearrange columns
all_data %<>%
  dplyr::select(-latitude,-longitude) %>%
  dplyr::select(id, clim_year, clim_month, clim_date, everything())

# adding in data that was originally missing due to bad gps coordinate format 
missing_data <- read_csv("Input_Files/missing_redo_1901-2024M.csv") %>%
  dplyr::select(-id2,-Elevation) %>%
  # Make data tall (one row per variable/site/month/year)
  gather(variable, value, -Year, -id1, -Latitude, -Longitude) %>%
  # Split month from variable name
  separate(variable, -2, into = c("variable", "month"), convert = TRUE) %>%
  # Spread again so that there is one row per site/year/month, one column per variable type
  spread(variable, value) %>%
  # Make a climate date column
  mutate(Year = as.numeric(Year),
         month = as.numeric(month),
         month_pad = str_pad(month, 2, pad = "0"),
         clim_date = as.numeric(paste0(Year, month_pad))) %>%
  # Drop unnecessary columns
  select(-month_pad) %>%
  # Rename to standard variable names
  rename(id = id1, longitude = Longitude, latitude = Latitude, clim_year = Year, clim_month = month, ppt_mm = PPT, srad = rsds, tmin = Tmin, tave = Tave, tmax = Tmax, CMD_mm = CMD, CMI_mm = CMI, chillDD = DD_0_, heatDD = DD_18_, coolDD = DD18_, growDD= DD5_, FFD = NFFD)

# Rearrange columns
missing_data %<>%
  dplyr::select(-latitude,-longitude) %>%
  dplyr::select(id, clim_year, clim_month, clim_date, everything())

# Add missing data back into main df
all_data <- all_data %>% 
  rows_update(missing_data, by = c("id", "clim_date"))

# Now save this data
write.csv(all_data, "Output_Files/climateNA_full_data_tall.csv", row.names = FALSE)

```

## Subset Climate Data

### 1 Year Timeframe

```{r}
#| label: subsetting 1 year climate data

# load the data
clim <- read.csv("Output_Files/climateNA_full_data_tall.csv", header = T)
specimens <- read.csv("Input_Files/FullSpecimenCCH2Data.csv", header = T)

#Note: Will eventually need to combine back in at least phenology scoring data(maybe not coords, etc from CCH2). This could likely best be done using the full entire class dataset from earlier code that is within Raphanus Phenology markdown


#subset to keep just the columns we need 
specimens <- specimens %>% dplyr::select(id, year, month, day)


#WUsing the full year (12 months) of data starting with the day and month of collection, but in the previous year, and ending with the date of collection. 

# METHODS: First, we need to calculate the date range. We will only include the month of collection in the year of collection if it was collected after the first week, so on the 8th or later. If it was collected before the 8th (1-7th), we do NOT include that month for the year of collection, and instead begin from the full month completed prior to the specimen's collection Additionally, if a specimen did not have a specific date of collection listed (n = 17), whcih displays as day=0 in the data frame, we do not feel confident in assuming that we should use said listed month as the end date, so the specimen will be treated the same as a specimen collected on the 1st-7th of the month. 


# Step 1: Identify the start and end month for the "climate year" for each specimen within the specimen df.These new columns should be in the clim_date format with year and month together (e.g. 202412 for December 2024).

# Step 1a: Identify the end month (this will then define the start month as 11 mo before). The newly coded end_mo column should contain the month of collection ( = month) if the day of collection column has a value >7. Otherwise, the end_mo column is assigned the month previous to collection (month -1). Special care was taken to address the cyclical nature of the calendar year, where Janaury is preceded by December.

specimens <- specimens %>% mutate(
  end_mo = if_else(day > 7, month, 
if_else(month == 1, 12, month - 1))) 

# Step 1b: Identify the start month and create its column; this will simply be the end month + 1 in order to capture the full 12 month (1 year) cycle. Once again, special care was taken to address the cyclical nature of the calendar year, where Janaury is preceded by December.


specimens <- specimens %>% mutate(start_mo = if_else(end_mo != 12, end_mo + 1, 1))

# Step 1c: Now there are two new columns in the specimens df, one each for start and end month. The year of collection must be appended to the month in order to get the start date and end date pertaining to each specimen that will narrow down the ClimateNA data. Special precautions were included to address the End Date scenario where a specimen collected before January 8th is assigned December as an end date, but the year must be listed (for both start and end date), as the year prior. Additionally, when constructing the Start Date character string, special conditions were added for specimens collected past December 7th, where, unlike any other collection date, the Start and End Date fall in the same year. 


#For End Date:
specimens <- specimens %>% mutate(end_date = if_else(end_mo == 12 & month == 1, paste0(year-1, str_pad(end_mo, 2, pad = "0")), paste0(year, str_pad(end_mo, 2, pad = "0"))), end_mo = as.integer(end_mo))

#For Start Date:
specimens <- specimens %>% mutate(start_date = if_else(month == 12 & end_mo == 12, paste0(year, str_pad(start_mo, 2, pad = "0")), paste0(year -1, str_pad(start_mo, 2, pad = "0"))), start_mo = as.integer(start_mo))

# Step 2: Join the specimens and clim dataframes to combine the newly adjusted CCH2 Data with the unfiltered ClimateNA data frame.

big_comb <- left_join(clim, specimens, by = "id")

# Step 3: To subset the climate data, keep only the rows for each specimen ID where the climate date is greater than or equal to the start date and less than or equal to the end date.

filter_comb <- big_comb %>% filter(clim_date >= start_date, clim_date <= end_date)

# Step 4: Convert the long/tall format climate data to wide format where you only have 1 row per individual with a column for each month and each climate variable that the specimen experienced over its year of interest (i.e. Tave_01 is the Average Temperature in January).


# Simplify to keep only necessary columns

filter_comb <- filter_comb %>% 
  select(-clim_year, -clim_date, -end_mo, -start_mo)

#Pivot the Data Frame to be wider and include new columns for every climate month paired with each climate variable.

wide_final_clim_subset <- filter_comb %>% pivot_wider(
    names_from = clim_month,
    values_from = c("CMD_mm", "CMI_mm", "chillDD", "heatDD", "coolDD", 
"growDD", "Eref", "FFD", "PAS", "ppt_mm", "RH", "srad", "tave", "tmax", "tmin")
  )


# Step 5: Write out the combined project csv... not going to write until I hopefully figure out the weird missing data

write.csv(wide_final_clim_subset, "Output_Files/climateNA_subset_data_tall.csv", row.names = FALSE)

```

#### Join and clean data

```{r}
#| label: join and clean

# read in the data
oneyr_clim <- read.csv("Output_Files/climateNA_subset_data_tall.csv")
fulldata<- read.csv("Output_Files/completed_specimen_data.csv")

# clean the county names in the phenology data
fulldata$county <- gsub("\\s*[Cc]ounty\\s*$", "", fulldata$county) # remove "county"
fulldata$county <- trimws(fulldata$county) # trim whitespace

# combine with phenology data
phen_1yr <- fulldata %>% 
  select(id, institutionCode, specificEpithet, year, month, day, county, decimalLatitude, decimalLongitude, Bud.Cluster, Flower, Fruit) %>% # select only necessary cols
  full_join(oneyr_clim) # join with climate data
```

#### PCA of climate

```{r}

# change -9999 to NA
phen_1yr[phen_1yr == -9999] <- NA

# calculate some reasonable per year/individual values from the monthly data (NATALIE - please look up what yearly variables are available through climateNA and write out how we would calculate each with our data - assuming it is a variable I didn't calculate below)

clim_1yr <- phen_1yr %>%
  mutate(
    # Summations
    sum_heatDD = rowSums(across(starts_with("heatDD_")), na.rm = TRUE),
    sum_coolDD = rowSums(across(starts_with("coolDD_")), na.rm = TRUE),
    sum_chillDD = rowSums(across(starts_with("chillDD_")), na.rm = TRUE),
    sum_growDD = rowSums(across(starts_with("growDD_")), na.rm = TRUE),
    sum_FFD    = rowSums(across(starts_with("FFD_")), na.rm = TRUE), # frost-free days
    sum_PAS    = rowSums(across(starts_with("PAS_")), na.rm = TRUE), # snow
    sum_ppt_mm = rowSums(across(starts_with("ppt_mm_")), na.rm = TRUE), 

    # Averages
    ave_CMD = rowMeans(across(starts_with("CMD_")), na.rm = TRUE), # clim moisture def
    ave_CMI = rowMeans(across(starts_with("CMI_")), na.rm = TRUE), # clim moisture index
    ave_Eref = rowMeans(across(starts_with("Eref_")), na.rm = TRUE), # evaporation
    ave_RH   = rowMeans(across(starts_with("RH_")), na.rm = TRUE), # relative humidity
    ave_srad = rowMeans(across(starts_with("srad_")), na.rm = TRUE), # solar rad
    ave_tave = rowMeans(across(starts_with("tave_")), na.rm = TRUE),

    # Hot/cold extremes (row-wise)
    max_tmax = do.call(pmax, c(across(starts_with("tmax_")), na.rm = TRUE)),
    min_tmin = do.call(pmin, c(across(starts_with("tmin_")), na.rm = TRUE))
  )

# id individuals with missing climate data (NATALIE: figure out why these are missing data - was it missing from the original climate data file, or did we lose it along the way?)
# rows missing data:
# bad_rows <- which(!complete.cases(clim_1yr) | rowSums(!is.finite(as.matrix(clim_1yr))) > 0) # 21 indivs without climate data
# remove bad rows
clim_1yr_clean <- clim_1yr[complete.cases(clim_1yr), ]

# subset for jsut climate data to run prcomp
clim_only_1yr_clean <- clim_1yr_clean %>% 
  select(sum_heatDD, sum_coolDD, sum_growDD, sum_FFD, sum_PAS, sum_ppt_mm, ave_CMD, ave_CMI, sum_chillDD, ave_Eref, ave_RH, ave_srad, ave_tave, max_tmax, min_tmin)

# calculate principal components
clim_pca <- prcomp(clim_only_1yr_clean, center = TRUE, scale = TRUE)

# display principal components
clim_pca$rotation

# Basic biplot
biplot(clim_pca, scale = 0)

# Extract scores and loadings
pca_scores <- as.data.frame(clim_pca$x)
loadings <- as.data.frame(clim_pca$rotation)

# Scale arrows for plotting
arrow_scale <- 11  # adjust to make arrows visible
library(scales)
library(cowplot)
library(maps)
ggplot(pca_scores, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.7) +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1 * arrow_scale, yend = PC2 * arrow_scale),
               arrow = arrow(length = unit(0.5, "cm")),
               color = "#E38471", size = 0.7) +
  geom_text(data = loadings,
            aes(x = PC1 * arrow_scale * 1.1, y = PC2 * arrow_scale * 1.1, label = rownames(loadings)),
            color = "#E38471", size = 3) +
  theme_minimal() +
  labs(x = "PC1 (45.3%)", y = "PC2 (22.0%)",
       title = "1yr climate PCA Biplot") +
  coord_equal()

# add back location and other info
phen_w_clim_pca_scores <- bind_cols(clim_1yr_clean, pca_scores)

pca_scores_forbiplot <- phen_w_clim_pca_scores %>%
  mutate(
    geo_color = rgb(
      rescale(decimalLongitude, to = c(0, 1)),        # red channel
      rescale(decimalLatitude,  to = c(0, 1)),        # green channel
      rescale(1 - abs(decimalLatitude - mean(decimalLatitude)) / max(decimalLatitude), to = c(0, 1)),  # blue channel
      alpha = 0.8
    )
  )

# Biplot with geographic coloring
# 1️⃣ Make the main PCA biplot
p_biplot <- ggplot(pca_scores_forbiplot, aes(x = PC1, y = PC2)) +
  geom_point(aes(color = geo_color), alpha = 0.8, size = 2) +
  geom_segment(data = loadings,
               aes(x = 0, y = 0,
                   xend = PC1 * arrow_scale,
                   yend = PC2 * arrow_scale),
               arrow = arrow(length = unit(0.5, "cm")),
               color = "black", size = 0.7) +
  geom_text(data = loadings,
            aes(x = PC1 * arrow_scale * 1.1,
                y = PC2 * arrow_scale * 1.1,
                label = rownames(loadings)),
            color = "black", size = 3) +
  theme_minimal() +
  labs(
    x = "PC1 (45.3%)",
    y = "PC2 (22.0%)",
    title = "1yr Climate PCA Biplot"
  ) +
  coord_equal() +
  theme(legend.position = "none")

# 2️⃣ Make a simple California outline map
california_map <- map_data("state") %>%
  dplyr::filter(region == "california")

p_map <- ggplot() +
  geom_polygon(data = california_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50", linewidth = 0.3) +
  geom_point(data = pca_scores_forbiplot,
             aes(x = decimalLongitude, y = decimalLatitude, color = geo_color),
             size = 1.6, alpha = 0.9) +
  coord_fixed(1.3, xlim = c(-125, -113), ylim = c(32, 43)) +
  theme_void() +
  theme(legend.position = "none")

# 3️⃣ Combine the two plots with the map inset
final_plot <- ggdraw() +
  draw_plot(p_biplot) +
  draw_plot(p_map, x = 0.01, y = 0.04, width = 0.6, height = 0.6) # adjust position/size

final_plot


### bilot of only flowering individuals colored by month

# Create reversed color palette
spectral_colors <- rev(brewer.pal(11, "Spectral")) # reversed palette (11 colors)
# Interpolate to 12 colors for all months
spectral_colors_12 <- colorRampPalette(spectral_colors)(12)

arrow_scale <- 18  # adjust to make arrows visible
# Plot
mo_biplot <- ggplot(pca_scores_forbiplot %>% filter(Flower != 0),
       aes(x = PC1, y = PC2)) +
  geom_point(aes(color = factor(month, levels = 1:12)), alpha = 0.8, size = 2) +
  geom_segment(data = loadings,
               aes(x = 0, y = 0,
                   xend = PC1 * arrow_scale,
                   yend = PC2 * arrow_scale),
               arrow = arrow(length = unit(0.5, "cm")),
               color = "black", size = 0.7) +
  geom_text(data = loadings,
            aes(x = PC1 * arrow_scale * 1.1,
                y = PC2 * arrow_scale * 1.1,
                label = rownames(loadings)),
            color = "black", size = 3) +
  scale_color_manual(values = spectral_colors_12,
                     labels = month.abb) +
  theme_minimal() +
  labs(
    x = "PC1 (45.3%)",
    y = "PC2 (22.0%)",
    title = "1yr Climate PCA Biplot\ncolor by month",
    color = "Month"
  ) +
  coord_equal() +
  theme(legend.position = "right")

# Calculate the variance explained by each principal component
per_variance_explained <- (clim_pca$sdev^2 / sum(clim_pca$sdev^2))*100
# Create a data frame for the variance explained
variance_df <- data.frame(
  PC = paste0("PC", 1:length(per_variance_explained)),
  Variance_Explained = per_variance_explained,
  Cumulative_Variance = cumsum(per_variance_explained)
)
variance_df #PC3 = 13.33% variance

# look at PC1 vals to see if any bimodality
library(ggplot2)
ggplot(pca_scores, aes(x = PC1)) +
  geom_histogram(bins = 100, fill = "skyblue", color = "white") +
  scale_x_continuous(breaks = pretty(pca_scores$PC1, n = 15)) +
  theme_minimal() +
  labs(title = "PC1 Distribution", x = "PC1 Score", y = "Count")

# make simpler df to work with
phen_sub <- phen_w_clim_pca_scores %>% 
  select(id,institutionCode,specificEpithet,year,month,day,county,decimalLatitude,decimalLongitude,Bud.Cluster,Flower,Fruit,PC1,PC2,PC3)

# Make PCA - colored by county for now
ggplot(phen_sub, aes(x=PC1, y=PC2, colour = county)) +
  geom_point() +
  labs(title = "Initial PCA", x = "PC1 (45.3%)", y = "PC2 (22.0%)")


# PICKUP could save this but looks like we should add more variables similar to what I had before? or draw diagonal ovals to group? TRY same PCA but colored by month collected:
# PICKUP add biplot back in!

ggplot(phen_sub, aes(x = PC1, y = PC2, colour = month)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_colour_gradientn(
    colours = c(
      "#313695",  # deep blue (winter)
      "#4575b4",  # blue (late winter)
      "#74add1",  # blue-green (early spring)
      "#abd9e9",  # greenish-blue (mid-spring)
      "#fdae61",  # orange (summer)
      "#f46d43",  # red-orange (late summer)
      "#d73027",  # red (fall)
      "#a50026",  # dark red (late fall)
      "#313695"   # loop back to deep blue
    ),
    limits = c(1, 12),
    breaks = 1:12,
    labels = month.abb
  ) +
  labs(
    title = "Seasonal Color Scale",
    x = "PC1 (45.3%)", 
    y = "PC2 (22.0%)",
    colour = "Month"
  ) +
  theme_minimal()
# seems like no really clear pattern... not a lot of hope to see the diff life cycles...
# maybe low vals of PC1 and high of PC1 show less flowering in Summer...? 
# From the biplot, I think that would be the more mild, more growing DD, but also less rain and hotter?? Maybe need to remove mountain individuals or whatever that tail to the bottom right is to get better PC axes for analysis

# split by -4 anyway and plot flowering histogram:
# Split data and filter for Flower != 0
phen_sub_filtered <- phen_sub %>%
  filter(Flower != 0) %>%                  # keep only non-zero Flower
  mutate(nor_coast_group = if_else(PC1 > 5 & PC2 > -2.5,
                         "Nor coast (PC1 > 5 & PC2 > -2.5)",
                         "Not nor coast"))  # create group

# THIS DOESN"T work... not sure how to divide so I separate out the 2 distributions... or if I can. 

# Histogram of month faceted by PC1 group
ggplot(phen_sub_filtered, aes(x = month)) +   # change x = month if preferred
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  facet_wrap(~ nor_coast_group) +
  labs(title = "Distribution of Flowering Dates by nor coast or not",
       x = "Day of Year",
       y = "Count") +
  theme_minimal()

# Look at distributions by county
# Filter for flowering specimens and create PC1 group
library(dplyr)
library(ggplot2)

# Filter for flowering specimens and only counties with ≥20 individuals
phen_sub_filtered <- phen_sub %>%
  filter(Flower != 0) %>%          # keep only flowering specimens
  add_count(county) %>%            # count individuals per county
  filter(n >= 20) %>%              # keep only counties with 20 or more
  select(-n)                        # remove helper column

# Plot: histogram per county
ggplot(phen_sub_filtered, aes(x = month)) +
  geom_histogram(binwidth = 1, fill = "pink", color = "black", boundary = 0) +
  facet_wrap(~ county, scales = "free_y") +
  scale_x_continuous(breaks = 1:12, labels = 1:12, limits = c(1, 12)) +
  labs(title = "Flowering Specimen Frequency by County",
       x = "Month",
       y = "Count") +
  theme_minimal() +
  theme(strip.text = element_text(size = 10))
```

#### Interactive plot color by PC1 from above to understand climate distribution

```{r}
# plot interactive map w points colored by collection month
      
specimen_map_leaflet <- leaflet(phen_sub) %>%
  addTiles() %>%  # Add OpenStreetMap tiles
  addCircleMarkers(
    lat = ~decimalLatitude, 
    lng = ~decimalLongitude, 
    radius = 4,  # Size of the markers
    fillColor = ~colorNumeric("Spectral", phen_sub$month)(month),
    color = "black",           # thin black border
    weight = 1,              # border thickness
    stroke = TRUE,             # show border
    fillOpacity = 1,
    popup = ~paste(
      "Scientific Name:", specificEpithet, "<br>",
      "Year:", year, "<br>",
      "Herbarium:", institutionCode
    ),
    label = ~paste(specificEpithet, institutionCode, id)
  ) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6) %>%
  addLegend(
    position = "bottomright",
    pal = colorNumeric("Spectral", phen_sub$month),
    values = ~month,
    title = "Month",
    opacity = 1
  )

saveWidget(specimen_map_leaflet, "maps/interactive_CA_specimen_by_month_all_Cal_phenology.html")

# Display the interactive plot
#interactive_plot2

# color by PC1
# make a custom palette
custom_palette <- colorRampPalette(rev(c(
  "#542788",  # purple
  "#053061",  # deep navy blue
  "#2166AC",  # strong blue
  "#4393C3",  # medium-light blue
  "#92C5DE",  # pale blue
  "#D9C48F",  # darker gray/yellow midpoint
  "#F4A582",  # light salmon
  "#D6604D",  # orange-red
  "#B2182B",  # deep red
  "#67000D"   # near-black red
)))
pal <- colorNumeric(palette = custom_palette(100), domain = phen_sub$PC1)

# Spectral is good but the yellow in the middle is hard to see

#PICKUP - finish picking pallette and fix color legend; 2 extreme PC1 vals driving things, maybe make red extreme black or red black too!
  
# build the map colored by PC1
specimen_map_leaflet <- leaflet(phen_sub) %>%
specimen_map_leaflet <- leaflet(phen_sub) %>%
  addTiles() %>%
  addCircleMarkers(
    lat = ~decimalLatitude, 
    lng = ~decimalLongitude, 
    radius = 4,  # Size of the markers
    fillColor = ~pal(PC1),    # fill uses the palette
    color = "black",           # thin black border
    weight = 0.5,              # border thickness
    stroke = TRUE,             # show border
    fillOpacity = 1,
    popup = ~paste(
      "Scientific Name:", specificEpithet, "<br>",
      "Year:", year, "<br>",
      "Herbarium:", institutionCode
    ),
    label = ~paste(specificEpithet, institutionCode, id)
  ) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6) %>%
  addLegend(
    position = "bottomright",
    pal = pal,
    values = ~PC1,
    title = "PC1",
    opacity = 1
  )

saveWidget(specimen_map_leaflet, "maps/interactive_CA_specimen_by_PC1_all_Cal_phenology.html")


# color by PC2 (22% var)
custom_palette <- colorRampPalette(brewer.pal(11, "Spectral"))
pal <- colorNumeric(palette = custom_palette(100), domain = phen_sub$PC2)

specimen_map_leaflet <- leaflet(phen_sub) %>%
  addTiles() %>%
  addCircleMarkers(
    lat = ~decimalLatitude, 
    lng = ~decimalLongitude, 
    radius = 4,
    fillColor = ~pal(PC2),   # fill uses Spectral palette
    color = "black",         # thin black border
    weight = 0.5,            # border thickness
    stroke = TRUE,           # show border
    fillOpacity = 1,
    popup = ~paste(
      "Scientific Name:", specificEpithet, "<br>",
      "Year:", year, "<br>",
      "Herbarium:", institutionCode
    ),
    label = ~paste(specificEpithet, institutionCode, id)
  ) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6) %>%
  addLegend(
    position = "bottomright",
    pal = pal,
    values = ~PC2,
    title = "PC2",
    opacity = 1
  )

saveWidget(specimen_map_leaflet, "maps/interactive_CA_specimen_by_PC2_all_Cal_phenology.html")

# color by PC2
custom_palette <- colorRampPalette(brewer.pal(11, "Spectral"))
pal <- colorNumeric(palette = custom_palette(100), domain = phen_sub$PC2)

specimen_map_leaflet <- leaflet(phen_sub) %>%
  addTiles() %>%
  addCircleMarkers(
    lat = ~decimalLatitude, 
    lng = ~decimalLongitude, 
    radius = 4,
    fillColor = ~pal(PC2),   # fill uses Spectral palette
    color = "black",         # thin black border
    weight = 0.5,            # border thickness
    stroke = TRUE,           # show border
    fillOpacity = 1,
    popup = ~paste(
      "Scientific Name:", specificEpithet, "<br>",
      "Year:", year, "<br>",
      "Herbarium:", institutionCode
    ),
    label = ~paste(specificEpithet, institutionCode, id)
  ) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6) %>%
  addLegend(
    position = "bottomright",
    pal = pal,
    values = ~PC2,
    title = "PC2",
    opacity = 1
  )

saveWidget(specimen_map_leaflet, "maps/interactive_CA_specimen_by_PC2_all_Cal_phenology.html")

# investigate collection locations of flowering specimens collected in Sept-December (maybe through January would be wise..?) 

```

#### divide data plot phenology

```{r}
# Start with all data divided into pre 1950 and post 2000
# remove Sierran individuals (PC2 < -6)
phen_sub_no_sierra <- phen_sub %>% filter(PC2>-6)

# quick check how many specimens per year
hist(phen_sub_no_sierra$year, breaks=50)
# bulk of specimens actually between 1950 and 2000...

# divided data into groups by year
phen_sub_2_yr_groups <- phen_sub %>%
  filter(Flower != 0) %>%  # keep only non-zero Flower
  mutate(year_group = case_when(
    year <= 1950 ~ "Year ≤ 1950",
    year >= 2000 ~ "Year ≥ 2000",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(year_group))  # keep only the two groups of interest

# Histogram of flowering month, faceted by year group
ggplot(phen_sub_2_yr_groups, aes(x = month)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  facet_wrap(~ year_group) +
  labs(
    title = "Distribution of Flowering Dates: ≤1950 vs ≥2000",
    x = "Month",
    y = "Count"
  ) +
  theme_minimal()

# look at intervening years too:
phen_sub_3_yr_groups <- phen_sub %>%
  filter(Flower != 0) %>%  # keep only non-zero Flower
  mutate(year_group = case_when(
    year <= 1950 ~ "Year ≤ 1950",
    year > 1950 & year < 2000 ~ "1950–2000",
    year >= 2000 ~ "Year ≥ 2000",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(year_group)) %>%
  mutate(
    year_group = factor(
      year_group,
      levels = c("Year ≤ 1950", "1950–2000", "Year ≥ 2000")
    )
  )

# Histogram of flowering month, faceted by year group
ggplot(phen_sub_3_yr_groups, aes(x = month, fill = specificEpithet)) +
  geom_histogram(
    binwidth = 1,
    color = "black",
    position = "dodge"  # side-by-side bars for species
  ) +
  scale_x_continuous(
    breaks = 1:12,
    labels = 1:12,
    expand = c(0, 0.5)  # prevents cutting off month 12
  ) +
  scale_y_continuous(
    breaks = seq(0, max(table(phen_sub_filtered$month)), by = 5)
  ) +
  facet_wrap(~ year_group) +
  labs(
    title = "Distribution of Flowering Dates by Time Period and Species",
    x = "Month",
    y = "Count",
    fill = "Species"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 12),
    legend.position = "top")

# color by species


```

### 9-Month Timeframe

```{r}
#| label: subsetting 9 month climate data

# load the data
clim <- read.csv("Output_Files/climateNA_full_data_tall.csv", header = T)
specimens <- read.csv("Input_Files/FullSpecimenCCH2Data.csv", header = T)

#Note: Will eventually need to combine back in at least phenology scoring data(maybe not coords, etc from CCH2). This could likely best be done using the full entire class dataset from earlier code that is within Raphanus Phenology markdown


#subset to keep just the columns we need 
specimens <- specimens %>% dplyr::select(id, year, month, day)

library(dplyr)
library(zoo)
specimens <- specimens %>%
  mutate(
    # Construct a "yearmon" object from year+month
    ym = as.yearmon(paste(year, month), "%Y %m"),
    # If day <= 7, subtract one month, else keep the same
    end_ym = if_else(day > 7, ym, ym - 1/12),
    # Convert back to Date (first day of month)
    end_mo = as.Date(end_ym)
  )
```

### 5-Year Timeframe

```{r}
#| label: subsetting 5 year climate data

# load the data
clim <- read.csv("Output_Files/climateNA_full_data_tall.csv", header = T)
specimens <- read.csv("Input_Files/FullSpecimenCCH2Data.csv", header = T)

#Note: Will eventually need to combine back in at least phenology scoring data(maybe not coords, etc from CCH2). This could likely best be done using the full entire class dataset from earlier code that is within Raphanus Phenology markdown

#subset to keep just the columns we need 
specimens <- specimens %>% dplyr::select(id, year, month, day)

#Change clim_date column from integer to date

library(dplyr)
library(zoo)
specimens <- specimens %>%
  mutate(
    # Construct a "yearmon" object from year+month
    ym = as.yearmon(paste(year, month), "%Y %m"),
    # If day <= 7, subtract one month, else keep the same
    end_ym = if_else(day > 7, ym, ym - 1/12),
    # Convert back to Date (first day of month)
    end_mo = as.Date(end_ym)
  )
```

## Examine Climate Correlations

```{r}
#| label: climate correlations
# use pairs.panels() from the psych package to make a correlation plot for all the climate variables. This will show us if, for example, temperature and growing degree days are correlated (they should be!). this will be important information for later analyses

```

## Extra Code

#### Code used to resolve the issue with Missing Climate Variables

```{r}
#| label: Missing Climate Data Fix
#| eval: false
#| include: false

#The specimens that did not receive proper climate data downloads were listed as "-9999" for each climate variable. 

#Filter the list of specimens to only include the ones without proper climate data downloads using Climatic Moisture Deficit (January) as a filter, and create a list of those specimens
check_up <- wide_final_clim_subset %>% filter(CMD_mm_1 < 0) 
check_up <- as.list(check_up$id)

#Filter the CCH2 data frame (specimens) to only get the latitude and longitude for the messed up specimen IDs
bad_clim <- specimens %>% filter( id %in% check_up) %>% select(id, decimalLatitude, decimalLongitude)



#There are different dahses/hyphen characters present in longitude values that interfered with the ability to pull climate data from ClimateNA .

#Convert unusable dash characters to correct negative sign for longitude values
bad_clim$decimalLongitude <- gsub("−", "-", bad_clim$decimalLongitude)
bad_clim$decimalLongitude <- gsub("–", "-", bad_clim$decimalLongitude)

# Convert to numeric
bad_clim$decimalLongitude <- as.numeric(bad_clim$decimalLongitude)

#Write a CSV so the proper climate data can be downloaded
#write.csv(bad_clim, "Redo_Climate_Specimens.csv", row.names = FALSE)

```

#### Code from Bontrager that includes calculation of a Collection Day of Year (number of days since Sept. 1):

```{r}
#| label: bontrager DOWY code
#| eval: false
#| include: false

# First we need to calculate a consistent number representing collection date. For now, we will do it as the Bontrager paper did and use Sept 01 as the start of the day count, then the date of collection as the last date. Bontrager code to adapt for our data is below. We will obviously want to remove the stuff about species, use our data, and make sure our colnames match or change the code to match.

specimens_compiled = read_csv("rainfall-fitness/data/compiled_specimen_data.csv") %>% 
  mutate(species = str_c(genus, "_", species)) %>% 
  filter(!(species %in% c("s_barbatus", "s_cordatus", "s_morrisonii", "s_tortuosus"))) %>% 
  rename(collection_year = year, collection_day = day, collection_month = month) %>% 
  # mutate(collection_day_assumed = if_else(is.na(collection_day), 15, collection_day)) %>% 
  mutate(collection_date = make_date(year = collection_year, day = collection_day, month = collection_month)) %>% 
  mutate(sept_1_date = make_date(year = collection_year-1, month = 9, day = 1),
         collection_DOWY = as.numeric(collection_date - sept_1_date))
```

#### Code from bontrager for phenology analyses and figures (Bontrager et al 2025)

```{r}
#| label: bontrager phenology code
#| eval: false
#| include: false
# PHENOLOGY ----
# Q1B) Which components of climate influenced specimen phenology? ----------------

## Model 1: temperature and precipitation on pheno ----

mod_list1_p = vector(mode = "list", length = length(species_list))

for (i in 1:length(species_list)) {
  
  species_choose = species_list[i]
  
  print(i)
  print(species_choose) 
  
  dat_slim = dat1 %>% 
    filter(species == species_choose) %>% 
    mutate(re_group = as.integer(as.factor(group)))
  
  priors <- c(prior(normal(0, 1), class = b))
  # Use default student t prior on intercept
  
  mod = brm(weighted_fruits|trials(total_repro) ~ 
              collection_DOWY_scaled + 
              tave_rain_avg_scaled + 
              log_ppt_mm_rain_avg_scaled + 
              (1|re_group), 
            family = binomial(),
            prior = priors,
            cores = 2,
            iter = 4000,
            init = 0,
            data = dat_slim,
            refresh = 0,
            file = glue("rainfall-fitness/model_fits/pheno_model1_{species_choose}.rds"))
  
  print(summary(mod))
  
  mod_list1_p[[i]] = mod
  
}

names(mod_list1_p) = species_list

### Make table of parameters and rhats ----

m1_all_p = map_df(mod_list1_p, summarise_draws, .id="species", ~quantile(.x, probs = c(0.025, 0.975)), mean, median, sd, rhat, ess_bulk, ess_tail) %>% 
  left_join(species_names) %>% 
  rename(lower = `2.5%`, upper = `97.5%`) 

# Check Rhats for convergence 
# sum should be 0
sum(m1_all_p$rhat >= 1.01)
# All converge with 4000 iterations

# Make a slimmed down table for point-range plots

param_list_m1_p = m1_all_p %>% 
  dplyr::select(-species_nice, -species_phy) %>%
  filter(variable %in% c( "b_tave_rain_avg_scaled", "b_log_ppt_mm_rain_avg_scaled", "b_collection_DOWY_scaled")) %>% 
  mutate(variable_nice = fct_recode(variable,
                                    "Average temperature\n(first rainy month until species'\naverage collection month)" =
                                      "b_tave_rain_avg_scaled",
                                    "Total precipitation\n(first rainy month until species'\naverage collection month)" =
                                      "b_log_ppt_mm_rain_avg_scaled",
                                    "Collection DOWY" = "b_collection_DOWY_scaled"),
         sig = if_else((upper > 0 & lower > 0), "sig_pos",
                       if_else((upper <0 & lower < 0), "sig_neg", "ns"))) %>%
  mutate(response = "pheno", model = "model1", model_eq = "weighted_fruits|trials(total_repro) ~ collection_DOWY_scaled + tave_rain_avg_scaled + log_ppt_mm_rain_avg_scaled + (1|re_group)")


## Model 2: CWD ----

mod_list2_p = vector(mode = "list", length = length(species_list))

for (i in 1:length(species_list)) {
  
  species_choose = species_list[i]
  
  print(i)
  print(species_choose) 
  
  dat_slim = dat1 %>% 
    filter(species == species_choose) %>% 
    mutate(re_group = as.integer(as.factor(group)))
  
  priors <- c(prior(normal(0, 1), class = b))
  # Use default student t prior on intercept
  
  mod = brm(weighted_fruits|trials(total_repro) ~ 
              collection_DOWY_scaled + 
              cwd_rain_avg_scaled + 
              (1|re_group), 
            family = binomial(),
            prior = priors,
            cores = 2,
            iter = 4000,
            init = 0,
            data = dat_slim,
            refresh = 0,
            file = glue("rainfall-fitness/model_fits/pheno_model2_{species_choose}.rds"))
  
  print(summary(mod))
  
  mod_list2_p[[i]] = mod
  
}

names(mod_list2_p) = species_list


### Make table of parameters and rhats ----

m2_all_p = map_df(mod_list2_p, summarise_draws, .id="species", ~quantile(.x, probs = c(0.025, 0.975)), mean, median, sd, rhat, ess_bulk, ess_tail) %>% 
  left_join(species_names) %>% 
  rename(lower = `2.5%`, upper = `97.5%`)

# Check Rhats for convergence 
# sums should all be 0
sum(m2_all_p$rhat >= 1.01)
# all good but anceps ess are low

param_list_m2_p = m2_all_p %>% 
  dplyr::select(-species_nice, -species_phy) %>%
  filter(variable %in% c("b_cwd_rain_avg_scaled", "b_collection_DOWY_scaled")) %>% 
  mutate(variable_nice = fct_recode(variable,
                                    "Total climatic water deficit\n(first rainy month until species'\naverage collection month)" = "b_cwd_rain_avg_scaled",
                                    "Collection DOWY" = "b_collection_DOWY_scaled"),
         sig = if_else((upper > 0 & lower > 0), "sig_pos",
                       if_else((upper <0 & lower < 0), "sig_neg", "ns"))) %>%
  mutate(response = "pheno", model = "model2", model_eq = "weighted_fruits|trials(total_repro) ~ collection_DOWY_scaled + cwd_rain_avg_scaled + (1|re_group)")


# Figure 1: pheno vs. climate multipanel ----


plot_list1p_p = vector(mode = "list", length = length(species_list))

for (i in 1:length(species_list)) {
  
  species_choose = species_list[i]
  
  dat_slim = dat1 %>%
    filter(species == species_choose) %>%
    mutate(re_group = as.integer(as.factor(group)))
  
  plot_list1p_p[[i]] = mod_list1_p[[i]] %>%
    ggemmeans(., terms = c("log_ppt_mm_rain_avg_scaled [all]")) %>%
    # ggemmeans(., terms = c("log_ppt_mm_rain_avg_scaled [all]"), bias_correction = TRUE) %>% 
    as_tibble() %>%
    mutate(species = species_choose)
  
}

pheno_ppt_panel_plot = plot_list1p_p %>%
  bind_rows() %>%
  mutate(ppt_mm_rain_avg = exp(x * sd(dat1$log_ppt_mm_rain_avg) + mean(dat1$log_ppt_mm_rain_avg))) %>%
  mutate(variable = "b_log_ppt_mm_rain_avg_scaled") %>%
  left_join(., param_list_m1_p) %>% # add on significance to color lines
  left_join(., species_names) %>%
  ggplot() +
  geom_point(data = dat1, aes(x = ppt_mm_rain_avg, y = pheno_stage), alpha = 0.2) +
  geom_line(aes(x = ppt_mm_rain_avg, y = predicted, color = sig)) +
  geom_ribbon(aes(x = ppt_mm_rain_avg, ymin = conf.low, ymax = conf.high, fill = sig), alpha = 0.6) +
  labs(y = "Phenological stage") +
  scale_x_log10(name = "Precipitation (mm), log scale") +
  guides(color = "none", fill = "none")+
  scale_color_manual(values = c("grey60", "tomato", "dodgerblue")) +
  scale_fill_manual(values = c("grey60", "tomato", "dodgerblue")) +
  facet_wrap(. ~ species_nice, nrow = 3, scales = "free_y") +
  theme_cowplot() +
  theme(axis.title.x = element_text(hjust = 0.5), strip.text = element_text(face = "italic"), axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5)); pheno_ppt_panel_plot


ests1_p = ggplot(data = filter(param_list_m1_p, variable != "b_collection_DOWY_scaled")) +
  geom_point(aes(x = mean, y = species_phy_ordered, color = sig), size = 1) +
  geom_errorbarh(aes(xmin = lower, xmax = upper, y = species_phy_ordered, color = sig), height = 0, size = 0.7) +
  facet_grid(.~ variable_nice) +
  scale_color_manual(values = c("grey60", "tomato", "dodgerblue")) +
  scale_x_continuous(limits = c(-1, 1)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  guides(color = "none") +
  labs(y = "Species", x = "Estimated effect on\nphenology") +
  theme(axis.text.y = element_text(face = "italic", hjust = 0.5), axis.title.y = element_blank()); ests1_p

ests2_p = ggplot(data = filter(param_list_m2_p, variable != "b_collection_DOWY_scaled")) +
  geom_point(aes(x = mean, y = species_phy_ordered, color = sig), size = 1) +
  geom_errorbarh(aes(xmin = lower, xmax = upper, y = species_phy_ordered, color = sig), height = 0, size = 0.7) +
  facet_grid(.~ variable_nice) +
  scale_color_manual(values = c("grey60", "tomato", "dodgerblue")) +
  scale_x_continuous(limits = c(-1, 1)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  guides(color = "none") +
  labs(y = "Species", x = "Estimated effect on\nphenology") +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank()); ests2_p

p_centered = plot_grid(NULL, p, NULL, rel_heights = c(0.1, 1, 0.1), ncol = 1)

p_all_p = plot_grid(p_centered, ests1_p, NULL, ests2_p, rel_widths = c(0.3, 1, 0.05, 0.38), rel_heights = c(0.8, 1, 1), nrow = 1,
                    labels = c("", "b.", "c."), label_x = c(0, 0.25, 0)); p_all_p

plot_grid(pheno_ppt_panel_plot, p_all_p, rel_heights = c(0.8, 1, 1), nrow = 2,
          labels = c("a.", "", ""), label_x = c(0, 0.25, 0))

ggsave("rainfall-fitness/figs/Fig1_pheno_mod1_multipanel.pdf", height = 10, width = 9)
ggsave("rainfall-fitness/figs/Fig1_pheno_mod1_multipanel.png", height = 10, width = 9)



# Figure S8: temp and cwd on pheno panel plot ----

plot_list1t_p = vector(mode = "list", length = length(species_list))

for (i in 1:length(species_list)) {
  
  species_choose = species_list[i]
  
  dat_slim = dat1 %>%
    filter(species == species_choose) %>%
    mutate(re_group = as.integer(as.factor(group)))
  
  plot_list1t_p[[i]] = mod_list1_p[[i]] %>%
    ggemmeans(., terms = c("tave_rain_avg_scaled [all]")) %>%
    # ggemmeans(., terms = c("tave_rain_avg_scaled [all]", bias_correction = TRUE)) %>% 
    as_tibble() %>%
    mutate(species = species_choose)
  
}

pheno_temp_panel_plot = plot_list1t_p %>%
  bind_rows() %>%
  mutate(tave_rain_avg = x * sd(dat1$tave_rain_avg) + mean(dat1$tave_rain_avg)) %>%
  mutate(variable = "b_tave_rain_avg_scaled") %>%
  left_join(., param_list_m1_p) %>% # add on significance to color lines
  left_join(., species_names) %>%
  ggplot() +
  geom_point(data = dat1, aes(x = tave_rain_avg, y = pheno_stage), alpha = 0.2) +
  # geom_smooth(data = dat1, aes(x = tave_rain_avg, y = total_repro), method = "lm", alpha = 0.2) +
  geom_line(aes(x = tave_rain_avg, y = predicted, color = sig), size = 1) +
  geom_ribbon(aes(x = tave_rain_avg, ymin = conf.low, ymax = conf.high, fill = sig), alpha = 0.6) +
  guides(color = "none", fill = "none")+
  labs(y = "Phenological stage", x = "Average temperature (C)") +
  scale_color_manual(values = c("grey60", "tomato", "dodgerblue")) +
  scale_fill_manual(values = c("grey60", "tomato", "dodgerblue")) +
  facet_wrap(. ~ species_nice, nrow = 3, scales = "free_y") +
  theme_cowplot() +
  theme(axis.title.x = element_text(hjust = 0.5), strip.text = element_text(face = "italic"), axis.text.x = element_text(hjust = 0, vjust = 0.5)); pheno_temp_panel_plot


plot_list2_p = vector(mode = "list", length = length(species_list))

for (i in 1:length(species_list)) {

  species_choose = species_list[i]

  dat_slim = dat1 %>%
    filter(species == species_choose) %>%
    mutate(re_group = as.integer(as.factor(group)))

  plot_list2_p[[i]] = mod_list2_p[[i]] %>%
    ggemmeans(., terms = c("cwd_rain_avg_scaled [all]")) %>% 
    # ggemmeans(., terms = c("cwd_rain_avg_scaled [all]"), bias_correction = TRUE) %>% 
    as_tibble() %>%
    mutate(species = species_choose)

}

pheno_cwd_panel_plot = plot_list2_p %>%
  bind_rows() %>%
  mutate(cwd_rain_avg = x * sd(dat1$cwd_rain_avg) + mean(dat1$cwd_rain_avg)) %>%
  mutate(variable = "b_cwd_rain_avg_scaled") %>%
  filter(cwd_rain_avg > 0) %>%
  left_join(., param_list_m2_p) %>% # add on significance to color lines
  left_join(., species_names) %>%
  ggplot() +
  geom_point(data = dat1, aes(x = cwd_rain_avg, y = pheno_stage), alpha = 0.2) +
  geom_line(aes(x = cwd_rain_avg, y = predicted, color = sig)) +
  geom_ribbon(aes(x = cwd_rain_avg, ymin = conf.low, ymax = conf.high, fill = sig), alpha = 0.6) +
  labs(y = "Phenological stage", color = "Species", fill = "Species") +
  scale_x_continuous(name = "Climatic water deficit") +
  scale_color_manual(values = c("grey60", "tomato", "dodgerblue")) +
  scale_fill_manual(values = c("grey60", "tomato", "dodgerblue")) +
  guides(color = "none", fill = "none")+
  facet_wrap(. ~ species_nice, nrow = 3, scales = "free_y") +
  theme_cowplot() +
  theme(axis.title.x = element_text(hjust = 0.5), strip.text = element_text(face = "italic"), axis.text.x = element_text( hjust = 0, vjust = 0.5)); pheno_cwd_panel_plot


plot_grid(pheno_temp_panel_plot, pheno_cwd_panel_plot, rel_heights = c(1,1), nrow = 2,
          labels = c("a.", "b."))

ggsave("rainfall-fitness/figs/FigS8_pheno_mod12_cwd_temp_panel.pdf", height = 10, width = 9)
ggsave("rainfall-fitness/figs/FigS8_pheno_mod12_cwd_temp_panel.png", height = 10, width = 9)



# Q2B) Does later onset of seasonal rainfall, predicted to become more frequent under climate change, change phenology? ----

## Model 3: rainfall timing with recent dataset ----

mod_list3_p = vector(mode = "list", length = length(species_list))

for (i in 1:length(species_list)) {
  
  species_choose = species_list[i]
  
  print(i)
  print(species_choose) 
  
  dat_slim = dat3 %>% 
    filter(species == species_choose) %>% 
    mutate(re_group = as.integer(as.factor(group)))
  
  priors <- c(prior(normal(0, 1), class = b))
  # Use default student t prior on intercept
  
  
  mod = brm(weighted_fruits|trials(total_repro) ~ 
              collection_DOWY_scaled + 
              tave_rain_avg_scaled + 
              log_ppt_mm_rain_avg_scaled + 
              first_rain_DOWY_scaled + 
              (1|re_group),
            family = binomial(),
            prior = priors,
            cores = 2,
            iter = 4000,
            init = 0,
            data = dat_slim,
            refresh = 0,
            file = glue("rainfall-fitness/model_fits/pheno_model3_{species_choose}.rds"))
  
  print(summary(mod))
  
  mod_list3_p[[i]] = mod
  
}

names(mod_list3_p) = species_list


### Make table of parameters and rhats ----

m3_all_p = map_df(mod_list3_p, summarise_draws, .id="species", ~quantile(.x, probs = c(0.025, 0.975)), mean, median, sd, rhat, ess_bulk, ess_tail) %>% 
  left_join(species_names) %>% 
  rename(lower = `2.5%`, upper = `97.5%`)

# Check Rhats for convergence 
# sums should all be 0
sum(m3_all_p$rhat >= 1.01)

param_list_m3_p = m3_all_p %>% 
  dplyr::select(-species_nice, -species_phy) %>%
  filter(variable %in% c("b_collection_DOWY_scaled", "b_first_rain_DOWY_scaled", "b_tave_rain_avg_scaled", "b_log_ppt_mm_rain_avg_scaled")) %>% 
  mutate(variable_nice = fct_recode(variable,
                                    "First rain event DOWY" = "b_first_rain_DOWY_scaled",
                                    "Average temperature\n(first rainy month until species'\naverage collection month)" = "b_tave_rain_avg_scaled",
                                    "Total precipitation\n(first rainy month until species'\naverage collection month)" = "b_log_ppt_mm_rain_avg_scaled",
                                    "Collection DOWY" = "b_collection_DOWY_scaled"),
         sig = if_else((upper > 0 & lower > 0), "sig_pos",
                       if_else((upper <0 & lower < 0), "sig_neg", "ns"))) %>%
  mutate(response = "pheno", model = "model3", model_eq = "weighted_fruits|trials(total_repro) ~ collection_DOWY_scaled + tave_rain_avg_scaled + log_ppt_mm_rain_avg_scaled + first_rain_DOWY_scaled + (1|re_group)")


# Figure S9: point ranges model 3 on pheno ----

ests3_p = ggplot(data = param_list_m3_p) +
  geom_point(aes(x = mean, y = species_phy_ordered, color = sig), size = 1) +
  geom_errorbarh(aes(xmin = lower, xmax = upper, y = species_phy_ordered, color = sig), height = 0, size = 0.7) +
  facet_grid(.~ variable_nice) +
  scale_color_manual(values = c("grey60", "tomato", "dodgerblue")) +
  # scale_x_continuous(limits = c(-0.85, 0.85)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  guides(color = "none") +
  labs(y = "Species", x = "Estimated effect on\nphenology") +
  theme(axis.text.y = element_text(face = "italic", hjust = 0.5), axis.title.y = element_blank()); ests3_p

ggsave("rainfall-fitness/figs/FigS9_pheno_mod3_estimates.pdf", height = 6, width = 9)
ggsave("rainfall-fitness/figs/FigS9_pheno_mod3_estimates.png", height = 6, width = 9)


# Figure 2: first rain vs. pheno ----


plot_list3_p = vector(mode = "list", length = length(species_list))

for (i in 1:length(species_list)) {
  
  species_choose = species_list[i]
  
  dat_slim = dat3 %>%
    filter(species == species_choose) %>%
    mutate(re_group = as.integer(as.factor(group)))
  
  plot_list3_p[[i]] = mod_list3_p[[i]] %>%
    ggemmeans(., terms = c("first_rain_DOWY_scaled [all]")) %>% 
    # ggemmeans(., terms = c("first_rain_DOWY_scaled [all]"), bias_correction = TRUE) %>% 
    as_tibble() %>%
    mutate(species = species_choose)
  
}

pheno_germ_panel_plot = plot_list3_p %>%
  bind_rows() %>%
  mutate(first_rain_DOWY = x * sd(dat3$first_rain_DOWY) + mean(dat3$first_rain_DOWY)) %>%
  mutate(variable = "b_first_rain_DOWY_scaled") %>%
  left_join(., species_names) %>%
  left_join(., param_list_m3_p) %>% # add on significance to color lines
  ggplot() +
  geom_point(data = dat3, aes(x = first_rain_DOWY, y = pheno_stage), alpha = 0.2) +
  geom_line(aes(x = first_rain_DOWY, y = predicted, color = sig)) +
  geom_ribbon(aes(x = first_rain_DOWY, ymin = conf.low, ymax = conf.high, fill = sig), alpha = 0.6) +
  labs(y = "Phenological stage") +
  scale_x_continuous(name = "Timing of first rain event", breaks = c(0, 31, 62, 92, 123, 154, 182), labels = c("Sep 1", "Oct 1", "Nov 1", "Dec 1", "Jan 1", "Feb 1", "Mar 1")) +
  scale_color_manual(values = c("grey60", "tomato", "dodgerblue")) +
  scale_fill_manual(values = c("grey60", "tomato", "dodgerblue")) +
  guides(color = "none", fill = "none")+
  facet_wrap(. ~ species_nice, nrow = 3, scales = "free_y") +
  theme_cowplot() +
  theme(axis.title.x = element_text(hjust = 0.5), strip.text = element_text(face = "italic"), axis.text.x = element_text( hjust = 0, vjust = 0.5, angle = 90)); pheno_germ_panel_plot

# sept 30 = 30
# oct 31 = 61
# nov 30  = 91
# dec 31 = 122 
# jan 31 = 153
# feb 28 = 181
# march 31 = 212
# april 30 = 242
# may 31 = 273
# june 30 = 303
# July 31 = 334
# aug 31 = 365
# sept 30 = 395
# oct 31 = 426

ests3b_p = ggplot(data = filter(param_list_m3_p, variable == "b_first_rain_DOWY_scaled")) +
  geom_point(aes(x = mean, y = species_phy_ordered, color = sig), size = 1) +
  geom_errorbarh(aes(xmin = lower, xmax = upper, y = species_phy_ordered, color = sig), height = 0, size = 0.7) +
  facet_grid(.~ variable_nice) +
  scale_color_manual(values = c("grey60", "tomato", "dodgerblue")) +
  # scale_x_continuous(limits = c(-0.85, 0.85)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  guides(color = "none") +
  labs(y = "Species", x = "Estimated effect on\nphenology") +
  theme(axis.text.y = element_text(face = "italic", hjust = 0.5), axis.title.y = element_blank()); ests3b_p

p_centered = plot_grid(NULL, p, NULL, rel_heights = c(0.05, 1, 0.1), ncol = 1)

p1_p = plot_grid(p_centered, ests3b_p, rel_widths = c(0.3, 1), rel_heights = c(0.8, 1), nrow = 1, labels = c("", "a."), label_x = c(0, 0.25)); p1_p

pheno_germ_panel_plot_tall = plot_list3_p %>%
  bind_rows() %>%
  mutate(first_rain_DOWY = x * sd(dat3$first_rain_DOWY) + mean(dat3$first_rain_DOWY)) %>%
  mutate(variable = "b_first_rain_DOWY_scaled") %>%
  left_join(., species_names) %>%
  left_join(., param_list_m3_p) %>% # add on significance to color lines
  ggplot() +
  geom_point(data = dat3, aes(x = first_rain_DOWY, y = pheno_stage), alpha = 0.2) +
  geom_line(aes(x = first_rain_DOWY, y = predicted, color = sig)) +
  geom_ribbon(aes(x = first_rain_DOWY, ymin = conf.low, ymax = conf.high, fill = sig), alpha = 0.6) +
  labs(y = "Phenological stage") +
  scale_x_continuous(name = "Timing of first rain event", breaks = c(0, 31, 62, 92, 123, 154, 182), labels = c("Sep 1", "Oct 1", "Nov 1", "Dec 1", "Jan 1", "Feb 1", "Mar 1")) +
  scale_color_manual(values = c("grey60", "tomato", "dodgerblue")) +
  scale_fill_manual(values = c("grey60", "tomato", "dodgerblue")) +
  guides(color = "none", fill = "none")+
  facet_wrap(. ~ species_nice, nrow = 5, scales = "free_y") +
  theme_cowplot() +
  theme(axis.title.x = element_text(hjust = 0.5), strip.text = element_text(face = "italic"), axis.text.x = element_text( hjust = 0, vjust = 0.5, angle = 90)); pheno_germ_panel_plot_tall

p2_p = plot_grid(p1_p, pheno_germ_panel_plot_tall, rel_widths = c(0.8, 1), rel_heights = c(0.8, 1), nrow = 1, labels = c("", "b.")); p2_p

ggsave("rainfall-fitness/figs/Fig2_pheno_mod3_estimates_multipanel.pdf", height = 6, width = 12)
ggsave("rainfall-fitness/figs/Fig2_pheno_mod3_estimates_multipanel.png", height = 6, width = 12)
```

#### Code for pulling climateNA data with R - untested, but might work now by using the correctly formatted input file.

```{r}
# setwd("/Users/juliaharencar/Documents/Software/ClimateNA_v760 "); # set the ClimateNA root directory as the working directory 
# exe <- "/Users/juliaharencar/Documents/Software/ClimateNA_v760/ClimateNA_v7.60.exe" 
# inputFile = '/C:\\ InputFiles \\input_test.csv' 
# outputFile = '/C:\\test_normal.csv' 
# yearPeriod = '/C:\\ Perioddat \\Normal_1991_2020.nrm' 
# system2(exe,args= c('/Y', yearPeriod, inputFile, outputFile),wait=T) 
# dat <- read.csv(' C:/test_normal.csv'); head(dat) 
# 
# 
# 
# 
# # how I installed the package:
# devtools::install_github("burnett-m/climatenaR",build_vignettes = TRUE)
# library(climatenaR)
# 
# # try pulling data 
```

```{bash}
# use wine and rosetta with the 86_64 to call ClimateNA and download data
# didn't work before, might now with the fixed .csv file
# arch -x86_64 wine /Users/juliaharencar/Documents/Software/ClimateNA_v760/ClimateNA_v7.60.exe \
#   /Users/juliaharencar/Documents/Github/Raphanus_phenology/climateNA_locations.csv \
#   /Users/juliaharencar/Documents/Github/Raphanus_phenology/ClimateNA_outputt #Periods.txt
```
