---
title: "RaphanusPhenology"
author: "Natalie Veech"
format:
  html:
    code-fold: true
    code-tools: true
    code-overflow: wrap
    toc: true
    number-sections: true
theme:
  light: yeti
  dark: superhero
editor_options:
  chunk_output_type: console
editor: visual
---

# Methods

All specimen images were downloaded through the Consortium of California Herbaria (cch2.org). The search was filtered to only include specimens within *R. raphanistrum* and *R. sativus* that were collected within California and had corresponding. \~70 (**get exact number?!?!?!?**) specimens held in possession by the San Diego Natural History Museum were unimaged, but included in the full data set. These specimens were imaged using equipment (**more specificity here?**) from the University of California, Davis Center for Plant Diversity (Herbarium) and edited in Adobe Lightroom. The full specimen data set, prior to specimen exclusions, consisted of 1,251 specimens (\***this number may be wrong and not include SD**\*.)

```{r}
library(tidyverse)
```

## Specimen Exclusions

Many of the exclusions applied to the data were guided by the exclusion criteria described in the methods of Love and Mazer (2021). 1 specimen was excluded due to the blurriness of the image. 12 images were excluded for having a poor date range recorded for the time of collection. A date range was considered unusable if it spanned more than a month due to the frequently changing nature of *Raphanus* reproductive stages. Any specimens that were collected off of islands (e.g. Channel Islands) were excluded due to the unique climatic setting and lack of gene flow. 45 island specimens were removed from the data set based on this criterion. 2 specimens were deemed unfit for imaging, defined here as specimens that had an envelope containing plant matter on the specimen sheet, but no actual sample that could provide counts of reproductive structures. 71 specimens were excluded for having unreliable location data that was either difficult to assign coordinates to (see Assigning GPS Coordinates and Uncertainty) or too broad to draw accurate and precise climate data from. 5 specimens were excluded for having no listed collection date. 21 specimens were excluded on the basis of having poor metadata, lacking several pieces of information that are necessary understanding the spatial, temporal, and climatic context of the specimen. Lastly, 246 specimens were removed for possessing a spatiotemporal duplicate in the remaining data set, removing any repetition for a more cohesive data set.

## Assigning GPS Coordinates and Uncertainty

Any specimens without previously entered GPS coordinates were assigned GPS coordinates by hand. The locality data provided was used to narrow down a precise geographic position in Google Maps, occasionally with the assistance of ChatGPT or local records and articles to provide historical context. The coordinates were then taken from Google Maps and entered into a program that allows for visualizing uncertainty radii around a specified point. The uncertainty radius was hand-picked and cautiously selected to encompass the extent to which the locality description would logically still apply, similar to the approach described in the Georeferencing methodology in Bontrager et. al 2025. The coordinate system was for all hand-determined GPS coordinates was marked as World Geodetic System 1984 (WGS84), as used by both Google Maps and the map radius software (calcmaps.com).

WGS 84 was selected as the coordinate system to standardize all samples against due to its prevalence in location data following its introduction in 1987. Based on information gathered from NOAA, the one to two meter difference between coordinates in NAD83 and WGS84, the two coordinate systems are considered equal for the purposes of the Raphanus data. However, the differences between NAD27 and WGS84 are often greater, so any NAD27 points were converted to WGS84. The calculations were primarily conducted en masse by ChatGPT while spot checking using the calculator provided by the West Virginia Department of Environmental Protection. The variance between coordinates in NAD27 and WGS84 was on the scale of 10^-3^ or 10^-4^, correlating to a 10-100m discrepancy between the points.

Specimens with coordinates that were listed as NAD27 in the CCH2 data were converted to CCH2. There were two subsequent pathways for specimens with coordinates that were listed as "preprocessed" or "unknown", or were otherwise unlisted. For specimens collected prior to 1987, NAD27 was broadly assumed to be the coordinate system. For specimens collected after WGS84, it was assumed that any modern GPS devices, computers, or phones operate in WGS84. Due to the uncertainty associated with these assumptions, all original coordinate and geodetic datum data was saved separately.

An additional protocol was developed for examining and assigning missing uncertainty radii entries in the CCH2 data, all of which included coordinates that served as the starting judgement point. Based on the Love and Mazer (2021) protocol, 4000 meters was deemed the ceiling cutoff for assigning uncertainty radii at this stage. For specimens that had narrowly defined, "good" locality data, the specificity of location and inclusion of coordinates generated the assumption that the uncertainty was greater than 4000 meters and the specimen's uncertainty was labeled as 3999. If a specimen was determined to have broad, "bad" location data, the specimen label was then read to determine the origin of the coordinates. If the coordinates were present on the original label, it was assumed that the coordinates are accurate and that the location data was poorly recorded, earning the specimen an uncertainty radius assignment of 3999. However, if there were no coordinates present on the label, it was assumed that they were added at a later time based on the poor location and it cannot be safely assumed that the uncertainty radius is less than 4000 meters; these specimens were then assigned an uncertainty radius of 4001. The binomial nature of this distribution can eventually be reanalyzed and assigned a more specific uncertainty value depending on the nature of the climate data that is used.

By using the 111.1 kilometer \~ 1 degree conversion, it was decided that any specimen with an original latitudinal or longitudinal coordinate that was rounded to the nearest 0.1 degree (\~11 km) should be reexamined to determine whether the coordinates or uncertainty radii could be adjusted to more precisely fall on either side of the previously set 4000 meter uncertainty radius cutoff. Specimens with location descriptions that were determined to be broader in extent than 4000 meters, were assigned an uncertainty radius of 4001 if their previously listed uncertainty radius was not already greater than 4000. Specimens with location data that was deemed more specific than 4000 meters (e.g. parks, specific intersections, street addresses) then received adjusted, more precise coordinates in WGS84 and were reassigned a fitting uncertainty radius using the map radius software found at calcmaps.com.

## Choosing the Right Machine Learning Program

Many previously published scientific papers have also utilized images from various herbaria or citizen science platforms such as iNaturalist to study phenological trends. These papers were examined to get a sense of the software options available to annotate images, build models, and subsequently train machine learning programs to apply the model to a larger data set. There were many possible paths that varied in coding language, accessibility, complexity, and suitability for the needs of the project. Eventually, Roboflow was selected as the most practical choice for conducting the study, as the software provides the ease of completing annotation and training models in the same location. Additionally, Roboflow provides model evaluations with specific feedback such asprecision and recall metrics as well as confusion matrices to easily visualize errors in the model and coding prompts to deploy the model to other sets of images.

## Selecting Specimens for Model Training

It was initially decided that 100 specimens, slight greater than ten percent of the full data set, would be selected to be annotated and used to train the phenology models. The first consideration made was to include at least one specimen from each of the 31 herbaria that are represented by the complete specimen data set. Afterwards, specimens were added in to be relatively proportional to the distribution of the data set. When continuing to add in specimens after initial herbarium representation was accounted for, three factors received the majority of focus: time of collection, reproductive stage, and complexity of specimens. The time of collection was critical to creating a diverse, representative training set. While most specimens are collected during the spring and summer months, when flowering and fruiting is widespread, specimens collected during both the very beginning and the later months of the calendar year were fairly represent. Additionally, a majority of specimens possess both flowering and fruiting structures, so training set selection was carefully done to also include specimens that had not yet begun to fruit or that only had fruit left. Lastly, due to the many different conditional circumstances that can occur in herbarium specimens, such as image quality, age, human error, etc., both complex, overlapping specimens and well-defined clear cut specimens were selected for annotation to prepare the model for any unusual specimen appearances.

## Annotating Herbarium Specimens

**Insert text here about the procedures for annotating specimens and the subsequent additions of specimens to the training set**

```{bash}
#Insert Julia's code for resizing images, I think it was in bash?
```

## Training Model Versions

**Insert text here about how we picked different models to train, how each version did, the 70-20-10 distribution, model metrics and evaluations, etc.**

## Deploying Model from Roboflow

**Insert Julia's codes to get the JSONS and get JSONS to data frame**

```{bash}
#eventual code here
```

## Converting Model Data into a Usable Format

The following code was created to achieve two purposes. The first was to combine the JSON information on phenology (reproductive structure counts) after running both the original 155 annotated specimen images and the remaining 782 unannotated specimen images through the Roboflow model. The second was to combine all then phenology data with a subset of the data provided by CCH2. Intermediate steps were incorporated as needed to ensure compatibility when merging data frames.

```{r}
#Load tidyverse package into R from Library
library(tidyverse)

#Simplify CCH2 data down to the columns of interest
#Columns of interest were determined based on what would be necessary to identify each specimen and what would be necessary for using climate data to understand the effects of climate variables on flowering time
important_specimen_data <- read.csv("FullSpecimenCCH2Data.csv") %>% select(id, 
institutionCode, specificEpithet, recordedBy, eventDate, year:day, 
verbatimEventDate, county, locality, 
decimalLatitude:coordinateUncertaintyInMeters)

#Change "id" column from integer class to character class in CCH2 data frame
important_specimen_data$id <- as.character(important_specimen_data$id)

#Combine Annotated and Unannotated phenology class data into one big data frame
ann <- read.csv("annotated_inference_counts.csv")
unann <- read.csv("inference_counts_RETRY.csv")
entire_class_dataset <- bind_rows(ann, unann)

#Change the name of "image" column in phenology class data frame to "id" to align with CCH2 data frame
entire_class_dataset <- entire_class_dataset %>% rename(id = image)

#Images that were larger than Roboflow's maximum upload size of 20MB were resized via code and given the prefix "SML_" in the filename to discern the edited image. However, this prefix appeared in the id column of the phenology class data and needed to be removed in order to combine the ohenology class and CCH2 data frames
entire_class_dataset$id <- str_replace(entire_class_dataset$id, "^SML_", "")

#Remove any observation (e.g. duplicates) that is in phenology class, but not CCH2 data
rm <- setdiff(entire_class_dataset$id, important_specimen_data$id)
entire_class_dataset <- entire_class_dataset %>% filter(!entire_class_dataset$id 
                                                        %in% rm)

#Merge the CCH2 data frame and the full phenology class data frame into one large data frame, matching based on the id column with a precautionary full join
totalinfo <- full_join(entire_class_dataset, important_specimen_data, by = "id")

#Export the full combined data frame into a CSV for further uses
write.csv(totalinfo, "completed_specimen_data.csv", row.names = FALSE)

```

At a later time, it was discovered that, when running the code for the model on specimens, there was one previously hand-annotated specimen that was given counts of 0 bud clusters, 0 flowers, and 0 fruits. After hand checking this image, it was clear that all three reproductive structures could be found on the specimen. The original hand-annotation counts from Roboflow were assigned to the specimen using the following code:

```{r}
#Read in full complete data CSV
fulldata<- read.csv("completed_specimen_data.csv")


#Fix the specimen that was falsely determined to have no assigned reproductive structure counts by individually assigning the accurate counts from the original annotations done by hand in Roboflow
fulldata <- fulldata %>%
     mutate(
           Bud.Cluster = if_else(id == 1368028, 3, Bud.Cluster),
        Flower = if_else(id == 1368028, 2, Flower),
         Fruit = if_else(id == 1368028, 2, Fruit))

#Export this adjusted change back into the CSV
write.csv(fulldata, "completed_specimen_data.csv", row.names = FALSE)
```

## Extracting Climate Data

**Add in text here about how climate NA was Chosen and how we sourced data, lots of credit to Bontrager, etc.**

```{r}
#EVENTUAL BONTRAGER + Additional to us code here... unless we somehow link the entire other markdown into here if that's a thing
```

# Trial Plots

SEE BELOW NOTE UNDER ALL CHUNKS OF CODE: Instructions for additional phenology plots:\
Write the code for eahc plot in it's own chunk, and make sure the plots show up in the rendered html. (try both plots in same chunk vs separate chunks, and see which output html style you like better!)\

I THINK I DID THIS BELOW:\
1) First R chunk: Filter to only include specimens in flower, then create a histogram by month. SO the final plot should show the number of specimens collected in flower for a given month.

2\) Second R chunk: Do the same for Fruit

**Basic Frequency Plots:**

```{r}
#Bud Clusters
hist(fulldata$Bud.Cluster, breaks = 10, freq = TRUE, col = "darkorchid3", 
     border = "white", main = "Bud Cluster Frequency in California Raphanus 
     Herbarium Specimens", ylab = "Frequency", xlab = "# of Bud Clusters")
```

The majority of specimens had just one bud cluster, with a steep drop off between one and two bud clusters. While this is likely both a flaw in our model and just the inherently difficult-to-discern nature of bud clusters, other analyses could be attempted on bud cluster counts. In particular, we could ask for a representation of how far along specimens are in flowering by comparing a ratio of bud clusters to fruits for each specimen. This could assist in creating some sort of graph where we are able to see the proportion of flowers that were opened to the overall count of flowers, both opened and potential to have opened in the future. I could also benefit to doing additional research on bud cluster phenology to see what other questions could be asked, but this feels like a less reliable data point for the sake of our project.

```{r}
#Flowers
hist(fulldata$Flower, breaks = 40, freq = TRUE, col = "firebrick3", 
     border = "white", main = "Flower Frequency in California Raphanus 
     Herbarium Specimens", ylab = "Frequency", xlab = "# of Flowers")
```

There is a right skewed distribution for the flower frequency curve, with most observations falling between 0-10 flowers. I would like to find a way to use this data to further study latitudinal and longitudinal differences in flower count, as it is a factor that seems to have a significant effect on flowering times (Ridley & Ellstrand 2009), due to how much California's climate differs on both the E-W and N-S axis. Perhaps we could divide up latitudes into chunks and average over an area and/or use a continuous spectrum to see at what latitudes and what longitudes flower count peaks. (Not sure of the logistics of this). Additionally, fruiting to flowering frequency ratios could be interesting as well, and I'm still intrigued as to what analyses we might be able to do and what conclusions we could draw if we create a phenology scoring formula.

```{r}
#Fruits
hist(fulldata$Fruit, breaks = 25, freq = TRUE, col = "seagreen1", 
     border = "black", main = "Fruit Frequency in California Raphanus 
     Herbarium Specimens", ylab = "Frequency", xlab = "# of Fruits")
```

Similar to flowers, the distirbution curve for fruits is white skewing, with over 25% of the observations having between 0-5 fruits. I would like to repeat the above description of location based frequency analyses for fruits to see if there is a correlation of location frequency between fruits and flowers. Additionally, as mentioned earlier, fruiting to flowering frequency ratios could be interesting as well, and I'm still intrigued as to what analyses we might be able to do and what conclusions we could draw if we create a phenology scoring formula.

**Histograms for Frequency of Structure Per Month:**

```{r}
#Flowering and Months
flowering <- fulldata %>% filter(Flower > 0)
hist(flowering$month, breaks = seq(0, 12, 1), freq = TRUE, col = "firebrick3",
     border = "white", main = "Frequency of Specimens Collected While Flowering
     for Each Month", xlab = "Month", ylab = "Frequency")
```

By looking at this graph, we can tell that the majority of specimens that were flowering were collected in the Spring months between March and May. Over 250 flowering specimens were collected in April alone. When reading through the scientific articles, I was intrigued by the Ridley and Ellstrand 2009 papers, where they had 2 study sites for both southern inland and northern coastal specimens to be grown and studied. One specimen site was inland and southern, at UCR, and the other was coastal and northern, at Point Reyes. The authors mentioned that at the Point Reyes site, radishes tend to behave as a summer annual, flowering near the end of summer, while at the UCR site the Radish behave like winter annuals and flower in early spring. I think it could be interesting to expand on this histogram and incorporate location data to try and further understand this aforementioned trend and see if there is proof of this idea in the herbarium specimens.

```{r}
#Fruits and Months
fruiting <- fulldata %>% filter(Fruit > 0)
hist(fruiting$month, breaks = seq(0, 12, 1), freq = TRUE, col = "seagreen1",
     border = "black", main = "Frequency of Specimens Collected While Fruiting
     for Each Month", xlab = "Month", ylab = "Frequency")
```

A similar trend follows when looking at the graphs of fruit-month frequency and flower-month frequency with most specimens that were fruiting being collected in the spring, specifically April. Based on this observation, I first and foremost think it is important to go back and make a histogram of just when each specimen was collected within the year specifically, although based on the similarity of the two specified histograms, I can't imagine this would change much. On that note, I'd also like to make a histogram showing when specimens were collected across our entire range, since I actually couldn't necessarily narrow down off the cuff which decade had the most specimens collected. I'm not sure what sorts of questions would be more beneficially asked by fruits since much of the literature is on flowering time. Perhaps comparisons could be made for specimens collected within a year to see if there is a noticeable time difference between collection of early few flower specimens and late fruit specimens, but there is lots of subjectivity there due to the specimen just being a snapshot of any given population and possibilities for not enough data to pull this off. I'm also intrigued as to why for both flowers and fruits, there is a slight peak in October, but I don't know that there are enough specimens to parse that out.

\

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
