---
title: "RaphanusPhenology"
author: "Natalie Veech"
format:
  html:
    code-fold: true
    code-tools: true
    code-overflow: wrap
    toc: true
    number-sections: true
theme:
  light: yeti
  dark: superhero
editor_options:
  chunk_output_type: console
editor: visual
---

# US Methods

All specimens were downloaded from GBIF. After initial filtering on the GBIF website, 5,697 specimens were downloaded. The filers applied on the GBIF website included: Occurrence = Present, Basis of Record = Preserved Specimen, Scientific Name = "Raphanus \|L.", Country of Collection = U.S. Only. Specimens from all years and methods were included.

## Filtering GBIF Specimens

Filtering was conducted sequentially to exclude any specimens from the data set, whether it be from lack of metadata to reliably provide reproductive counts or capture climate data, or simple irrelevance to the geographic areas of interest for Raphanus.

The first method of filtering was to keep all specimens from states that fall into the three geographic regions of interest, and exclude all other states for the time being. There is the potential to revisit the inclusion of specimens from states of potential interest (i.e. Midwestern States, Hawaii) in the future. The geographic regions were separated based on map guidelines of climate regions as described by the Environmental Protection Agency, and thus, specimens in certain states may be divided up based on their location within multiple climate regions that span one state. The first geographic region, the West, consists of California, Oregon, and Washington. The second geographic region, North East, contains northern Virginia, Maryland, Delaware, New Jersey, Rhode Island, New York, Connecticut, Massachusetts, Vermont, New Hampshire, and Maine. The final region is the South East, which contains southern Virginia, portions of Texas, Louisiana, Alabama, Mississippi, Florida, Georgia, South Carolina, and North Carolina. Any specimen location that fell outside of our predefined regions was excluded through code. (n_state = #)

As a means to ensure the GBIF criteria were applied correctly when the download was first done, any specimens with a country code that was not used were excluded (n_country = #), and any specimens that had a Basis of Record that was not Preserved Specimen were excluded (n_basis = #).

Specimens without nay locality or VerbatimLocality information were also excluded due to the lack of metadata, especially due to the inability to identify coordinates for specimens without any previously listed. This constituted \_\_ specimens. (n_locality = #). Specimens with no locality information and just a county label were specified out separately later in the code.

Occurrence `mediaType` was not used as an exclusion filter; multimedia links were taken from `multimedia.txt` where available (including when `mediaType` was blank on the occurrence row).

A table was also created to view how many specimens were removed for each reason. Separate CSV files can be viewed for all removed specimens.

```{r}
#| label: Filtering GBIF Specimens based on set removal criteria
#| eval: false

# GBIF occurrence further filtering -------------------------------
# Set working directory to project root (Raphanus_Phenology) before sourcing.

library(dplyr)
library(readr)

# Paths (relative to project root)
path_occurrences <- "GBIF_occurrence_fixed.csv"
path_filtered_out <- "Output_Files/GBIF_occurrence_filtered.csv"
path_removed_county_only <- "Output_Files/GBIF_occurrence_removed_county_only.csv"
path_removed_mediaType_not_StillImage <- "Output_Files/GBIF_occurrence_removed_mediaType_not_StillImage.csv"
path_removed_not_preserved <- "Output_Files/GBIF_occurrence_removed_not_preserved_specimen.csv"
path_removed_cch2_catalog <- "Output_Files/GBIF_occurrence_removed_CCH2_catalog_overlap.csv"
path_removals_breakdown <- "Output_Files/GBIF_filter_removals_breakdown.csv"
path_cch2_catalog_ids <- "CCH2_2025_full_ID_list.txt"

# Load occurrence data (read as character to avoid parsing issues with messy GBIF fields)
occ <- read_csv(path_occurrences, col_types = cols(.default = col_character()), show_col_types = FALSE)

cch2_catalog_numbers <- read_csv(path_cch2_catalog_ids, show_col_types = FALSE) %>%
  pull(1) %>%
  as.character() %>%
  trimws() %>%
  unique()
cch2_catalog_numbers <- cch2_catalog_numbers[nzchar(cch2_catalog_numbers)]

# Define US states to exclude due to low specimen counts or irrelevance to geographic interests
states_exclude <- c(
  "Alaska (State)", "Arizona", "Arkansas", "Colorado",
  "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky",
  "Michigan", "Minnesota", "Missouri", "Montana", "Nebraska", "Nevada",
  "New Mexico", "North Dakota", "Ohio", "Oklahoma", "South Dakota",
  "Tennessee", "Utah", "West Virginia", "Wisconsin", "Wyoming", "0"
)

# Create output directory if needed
dir.create("Output_Files", showWarnings = FALSE)

# Apply filters sequentially to build removal breakdown

#n0 is representative of total starting row count, n_state keeps track of how many were removed for state, n_country drops any records where country code for collection isn't US, n_basis is for anything with a record that's not preserved specimen, 
n0 <- nrow(occ)
after_state <- occ %>% filter(!stateProvince %in% states_exclude)
n_state <- n0 - nrow(after_state)

after_country <- after_state %>% filter(countryCode == "US")
n_country <- nrow(after_state) - nrow(after_country)

after_basis <- after_country %>% filter(basisOfRecord == "PRESERVED_SPECIMEN")
n_basis <- nrow(after_country) - nrow(after_basis)

after_locality <- after_basis %>% filter(
  !( (is.na(locality) | trimws(locality) == "") &
     (is.na(verbatimLocality) | trimws(verbatimLocality) == "") )
)
n_locality <- nrow(after_basis) - nrow(after_locality)

# No mediaType / StillImage filter (specimens may still have rows in multimedia.txt)
after_media <- after_locality

# Optional export: rows where occurrence mediaType is not StillImage (diagnostic only)
mediaType_not_StillImage <- after_locality %>%
  filter(!grepl("StillImage", mediaType, fixed = TRUE))

# Drop specimens already represented in the CCH2 California download (catalogNumber match)
cat_trim <- function(x) ifelse(is.na(x), "", trimws(as.character(x)))
removed_cch2_catalog <- after_media %>%
  filter(cat_trim(catalogNumber) %in% cch2_catalog_numbers)
after_cch2 <- after_media %>%
  filter(!cat_trim(catalogNumber) %in% cch2_catalog_numbers)
n_cch2 <- nrow(after_media) - nrow(after_cch2)

occ_filtered <- after_cch2

# Breakdown table of removals (sequential)
removals_breakdown <- tibble(
  reason = c(
    "Excluded state (stateProvince in states_exclude)",
    "Country not US",
    "basisOfRecord not PRESERVED_SPECIMEN",
    "Empty locality (county-only)",
    "catalogNumber in CCH2_2025_full_ID_list.txt"
  ),
  n_removed = c(n_state, n_country, n_basis, n_locality, n_cch2)
) %>%
  mutate(
    n_remaining_after = n0 - cumsum(n_removed),
    .after = n_removed
  )

# Optional: print or view the table
print(removals_breakdown)

# Specimens with only county information (for potential county-level climate assignment)
county_only <- occ %>%
  filter(
    (is.na(locality) | trimws(locality) == "") &
    (is.na(verbatimLocality) | trimws(verbatimLocality) == "")
  )

# Specimens removed due to basisOfRecord not being PRESERVED_SPECIMEN
not_preserved_specimen <- occ %>% filter(basisOfRecord != "PRESERVED_SPECIMEN")

# Write breakdown table and removed specimens / filtered dataset
write_csv(removals_breakdown, path_removals_breakdown)
write_csv(not_preserved_specimen, path_removed_not_preserved)
write_csv(mediaType_not_StillImage, path_removed_mediaType_not_StillImage)
write_csv(county_only, path_removed_county_only)
write_csv(removed_cch2_catalog, path_removed_cch2_catalog)
write_csv(occ_filtered, path_filtered_out)


##Following code is for filtering the specimens that are in the GBIF Multimedia.txt File with most of same criteria that we applied to the full GBIF dataset:

# filter specimen image link document (original_gbif_download/multimedia.txt) for only specimens found in after_basis (include images for specimens without locality for now in case county is sufficient/has uniform climate)

#Read in the multimedia file with links
links <- read.delim(
  "original_gbif_download/multimedia.txt",
  header = TRUE,
  sep = "\t",
  quote = "\"",
  fill = TRUE,
  stringsAsFactors = FALSE
)

# One row per gbifID: merge duplicate multimedia rows by pasting unique non-empty
# values per column (e.g. several image URLs in `identifier`).
collapse_multimedia_by_gbifID <- function(df, sep = " | ") {
  collapse_col <- function(x) {
    x <- trimws(as.character(x))
    x <- x[!is.na(x) & x != ""]
    x <- unique(x)
    if (length(x) == 0) {
      return(NA_character_)
    }
    paste(x, collapse = sep)
  }
  df %>%
    group_by(gbifID) %>%
    summarise(across(everything(), collapse_col), .groups = "drop")
}

# Filter the Multimedia Links by linking the filtered full dataset to the multimedia textfile by the GBIF ID. 
after_basis$gbifID <- as.numeric(after_basis$gbifID)
links_subset <- links %>%
  semi_join(after_basis, by = "gbifID") %>%
  collapse_multimedia_by_gbifID()

# Print a subset with the filtered links into a table so that each specimen can be viewed as either a successful or unsuccessful pull of image
path_links_filtered <- "Output_Files/filtered_multimedia_260407.txt"
write.table(
  links_subset,
  file = path_links_filtered,
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)
```

```{r}
#| label: Filter original GBIF occurrence.txt (same criteria as fixed CSV)
#| eval: false

# Load raw Darwin Core occurrence table from the original GBIF download and apply
# the same sequential filters as `GBIF_occurrence_fixed.csv` (see chunk above).
library(dplyr)
library(readr)

path_occurrences_txt <- "original_gbif_download/occurrence.txt"
path_cch2_catalog_ids <- "CCH2_2025_full_ID_list.txt"
path_removals_breakdown_txt <- "Output_Files/GBIF_filter_removals_breakdown_occurrence_txt.csv"
path_links_filtered_from_occ_txt <- "Output_Files/filtered_multimedia_occurrence_txt.txt"
path_gbif_multimedia_diff <- "Output_Files/filtered_multimedia_gbifID_differences.csv"

occ_from_txt <- read_delim(
  path_occurrences_txt,
  delim = "\t",
  quote = "\"",
  col_types = cols(.default = col_character()),
  show_col_types = FALSE
)

cch2_catalog_numbers <- read_csv(path_cch2_catalog_ids, show_col_types = FALSE) %>%
  pull(1) %>%
  as.character() %>%
  trimws() %>%
  unique()
cch2_catalog_numbers <- cch2_catalog_numbers[nzchar(cch2_catalog_numbers)]

cat_trim <- function(x) ifelse(is.na(x), "", trimws(as.character(x)))

states_exclude <- c(
  "Alaska (State)", "Arizona", "Arkansas", "Colorado",
  "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky",
  "Michigan", "Minnesota", "Missouri", "Montana", "Nebraska", "Nevada",
  "New Mexico", "North Dakota", "Ohio", "Oklahoma", "South Dakota",
  "Tennessee", "Utah", "West Virginia", "Wisconsin", "Wyoming", "0"
)

dir.create("Output_Files", showWarnings = FALSE)

n0_txt <- nrow(occ_from_txt)
after_state_txt <- occ_from_txt %>% filter(!stateProvince %in% states_exclude)
n_state_txt <- n0_txt - nrow(after_state_txt)

after_country_txt <- after_state_txt %>% filter(countryCode == "US")
n_country_txt <- nrow(after_state_txt) - nrow(after_country_txt)

after_basis_txt <- after_country_txt %>% filter(basisOfRecord == "PRESERVED_SPECIMEN")
n_basis_txt <- nrow(after_country_txt) - nrow(after_basis_txt)

after_locality_txt <- after_basis_txt %>% filter(
  !( (is.na(locality) | trimws(locality) == "") &
     (is.na(verbatimLocality) | trimws(verbatimLocality) == "") )
)
n_locality_txt <- nrow(after_basis_txt) - nrow(after_locality_txt)

# No mediaType / StillImage filter (specimens may still have rows in multimedia.txt)
after_media_txt <- after_locality_txt

occ_filtered_from_cch2_catalog <- after_media_txt %>%
  filter(!cat_trim(catalogNumber) %in% cch2_catalog_numbers)
n_cch2_txt <- nrow(after_media_txt) - nrow(occ_filtered_from_cch2_catalog)

removals_breakdown_txt <- tibble(
  reason = c(
    "Excluded state (stateProvince in states_exclude)",
    "Country not US",
    "basisOfRecord not PRESERVED_SPECIMEN",
    "Empty locality (county-only)",
    "catalogNumber in CCH2_2025_full_ID_list.txt"
  ),
  n_removed = c(
    n_state_txt, n_country_txt, n_basis_txt, n_locality_txt,
    n_cch2_txt
  )
) %>%
  mutate(
    n_remaining_after = n0_txt - cumsum(n_removed),
    .after = n_removed
  )

print(removals_breakdown_txt)
write_csv(removals_breakdown_txt, path_removals_breakdown_txt)

# multimedia.txt: state, US, preserved, not CCH2 catalog overlap; no locality or
# mediaType requirement (cf. main chunk joining multimedia to after_basis).
occ_for_multimedia_from_txt <- after_basis_txt %>%
  filter(!cat_trim(catalogNumber) %in% cch2_catalog_numbers) %>%
  mutate(gbifID = as.numeric(gbifID))

# Same locality rule as elsewhere: TRUE if locality or verbatimLocality is non-empty
locality_flag_from_occ_txt <- occ_for_multimedia_from_txt %>%
  mutate(
    has_locality_occurrence_txt = !(
      (is.na(locality) | trimws(locality) == "") &
      (is.na(verbatimLocality) | trimws(verbatimLocality) == "")
    )
  ) %>%
  select(gbifID, has_locality_occurrence_txt) %>%
  distinct()

links_from_occ_txt <- read.delim(
  "original_gbif_download/multimedia.txt",
  header = TRUE,
  sep = "\t",
  quote = "\"",
  fill = TRUE,
  stringsAsFactors = FALSE
)

collapse_multimedia_by_gbifID <- function(df, sep = " | ") {
  collapse_col <- function(x) {
    x <- trimws(as.character(x))
    x <- x[!is.na(x) & x != ""]
    x <- unique(x)
    if (length(x) == 0) {
      return(NA_character_)
    }
    paste(x, collapse = sep)
  }
  df %>%
    group_by(gbifID) %>%
    summarise(across(everything(), collapse_col), .groups = "drop")
}

links_subset_from_occ_txt <- links_from_occ_txt %>%
  semi_join(occ_for_multimedia_from_txt, by = "gbifID") %>%
  collapse_multimedia_by_gbifID() %>%
  left_join(locality_flag_from_occ_txt, by = "gbifID")

write.table(
  links_subset_from_occ_txt,
  file = path_links_filtered_from_occ_txt,
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

# gbifIDs in one filtered multimedia list but not the other + locality from occurrence.txt
path_occurrences_fixed <- "GBIF_occurrence_fixed.csv"
occ_fixed <- read_csv(path_occurrences_fixed, col_types = cols(.default = col_character()), show_col_types = FALSE)
after_basis_fixed <- occ_fixed %>%
  filter(!stateProvince %in% states_exclude) %>%
  filter(countryCode == "US") %>%
  filter(basisOfRecord == "PRESERVED_SPECIMEN") %>%
  mutate(gbifID = as.numeric(gbifID))

ids_main_multimedia <- links_from_occ_txt %>%
  semi_join(after_basis_fixed, by = "gbifID") %>%
  collapse_multimedia_by_gbifID() %>%
  pull(gbifID)

ids_occ_multimedia <- links_subset_from_occ_txt %>% pull(gbifID)

locality_lookup_occ_txt <- occ_from_txt %>%
  mutate(
    gbifID = as.numeric(gbifID),
    has_locality_occurrence_txt = !(
      (is.na(locality) | trimws(locality) == "") &
      (is.na(verbatimLocality) | trimws(verbatimLocality) == "")
    )
  ) %>%
  group_by(gbifID) %>%
  summarise(has_locality_occurrence_txt = any(has_locality_occurrence_txt), .groups = "drop")

multimedia_gbifID_diff <- bind_rows(
  tibble(
    gbifID = as.numeric(setdiff(ids_main_multimedia, ids_occ_multimedia)),
    in_set = "only_filtered_multimedia_260407_logic"
  ),
  tibble(
    gbifID = as.numeric(setdiff(ids_occ_multimedia, ids_main_multimedia)),
    in_set = "only_filtered_multimedia_occurrence_txt_logic"
  )
) %>%
  left_join(locality_lookup_occ_txt, by = "gbifID") %>%
  arrange(in_set, gbifID)

write_csv(multimedia_gbifID_diff, path_gbif_multimedia_diff)
```

## Pulling Images from GBIF for Model Evaluation

```{python}
#| eval: false
#| label: Code to Pull Images from Direct Links in GBIF Multimedia File

import requests
import os
import csv

# Create folder
os.makedirs("gbif_images", exist_ok=True)

# Track failures
failed = []

# Start session (helps with some hosts)
session = requests.Session()
session.headers.update({
    "User-Agent": "Mozilla/5.0",
    "Referer": "https://www.gbif.org/"
})

# Load file
with open("../Output_Files/filtered_multimedia_260407.txt") as f:
    header = next(f).strip().split("\t")

    gbif_idx = header.index("gbifID")
    url_idx = header.index("identifier")

    rows = [line.strip().split("\t") for line in f if line.strip()]

# Download loop
for row in rows:
    try:
        gbif_id = row[gbif_idx]
        url = row[url_idx]

        if not url or url == "NA":
            continue

        url = url.strip('"')
        filename = f"gbif_images/{gbif_id}.jpg"

        if os.path.exists(filename):
            continue

        response = session.get(url, timeout=20)
        response.raise_for_status()

        with open(filename, "wb") as img_file:
            img_file.write(response.content)

        print(f"Downloaded {gbif_id}")

    except Exception as e:
        print(f"Failed {gbif_id}: {e}")
        failed.append({
            "gbifID": gbif_id,
            "url": url,
            "error": str(e)
        })

# ✅ Write failures to CSV
with open("failed_downloads.csv", "w", newline="") as f:
    writer = csv.DictWriter(f, fieldnames=["gbifID", "url", "error"])
    writer.writeheader()
    writer.writerows(failed)

print(f"\nFinished. {len(failed)} failures written to failed_downloads.csv")
```

## Getting US Counts and Metadata into usable format

```{python}
#| eval: false
#| label: Code to Turn GBIF JSONs to DF

import glob
import json
import pandas as pd
from collections import Counter

# Desired class order
class_order = ["Bud Cluster", "Flower", "Fruit"]

rows = []

# Loop through all JSON files
for file in glob.glob("*.json"):
    with open(file, "r") as f:
        result = json.load(f)
    
    # Extract class names from predictions
    classes = [pred["class"] for pred in result.get("predictions", [])]
    
    # Count how many of each class were detected
    counts = Counter(classes)
    
    # Image ID without extension
    image_id = file.replace(".json", "")
    
    # Build row dict with all classes in fixed order
    row = {"image": image_id}
    for cls in class_order:
        row[cls] = counts.get(cls, 0)
    
    rows.append(row)

# Build DataFrame with class order
df = pd.DataFrame(rows, columns=["image"] + class_order)

# Set image as index if desired (optional)
# df = df.set_index("image")

# Save to CSV
df.to_csv("gbif_repro_counts", index=False)

print("JSON data to gbif_repro_counts.csv")
print(df)

```

```{r}
#| label: Combining GBIF metaata with Roboflow Scores

#Load tidyverse package into R from Library
library(tidyverse)

#Simplify CCH2 data down to the columns of interest
#Columns of interest were determined based on what would be necessary to identify each specimen and what would be necessary for using climate data to understand the effects of climate variables on flowering time

//EDIT FOR GBIF VERSION: gbif_specimen_metadata <- read.csv("add gbif csv") %>% select(id, 
institutionCode, specificEpithet, recordedBy, eventDate, year:day, 
verbatimEventDate, county, locality, 
decimalLatitude:coordinateUncertaintyInMeters)

#Change "id" column from integer class to character class in CCH2 data frame
important_specimen_data$id <- as.character(gbif_specimen_metadata$id)

#Combine Annotated and Unannotated phenology class data into one big data frame
ann_gbif <- read.csv("add gbif csv")
unann <- read.csv("add gbif csv")
entire_gbif_class_dataset <- bind_rows(ann, unann)

#Change the name of "image" column in phenology class data frame to "id" to align with CCH2 data frame
entire_gbif_class_dataset <- entire_gbif_class_dataset %>% rename(id = image)

#Images that were larger than Roboflow's maximum upload size of 20MB were resized via code and given the prefix "SML_" in the filename to discern the edited image. However, this prefix appeared in the id column of the phenology class data and needed to be removed in order to combine the phenology class and CCH2 data frames.
entire_gbif_class_dataset$id <- str_replace(entire_gbif_class_dataset$id, "^SML_", "")

#Remove any observation (e.g. duplicates) that is in phenology class, but not CCH2 data
rm_gbif <- setdiff(entire_gbif_class_dataset$id, gbif_specimen_metadata$id)
entire_class_dataset <- entire_class_dataset %>% filter(!entire_class_dataset$id 
                                                        %in% rm)

#Merge the CCH2 data frame and the full phenology class data frame into one large data frame, matching based on the id column with a precautionary full join
totalinfo_gbif <- full_join(entire_gbif_class_dataset, gbif_specimen_metadata, by = "id")

#Export the full combined data frame into a CSV for further uses
write.csv(totalinfo_gbif, "Output_Files/completed_gbif_specimens.csv", row.names = FALSE)

```

## Basic Data Analysis

```{r}
library(dplyr)
library(ggplot2)
library(patchwork)
library(lubridate)
library(colorblindcheck)

gbif_full_flw<- read.csv("gbif_repro_counts_merged.csv") %>% mutate(monthname = month.abb[as.numeric(month)])%>% filter(Flower > 0)

west <- c("California", "Oregon", "Washington")

northeast <- c("Maryland", "Delaware", "New Jersey", "Rhode Island", "New York", "Connecticut", "Massachusetts", "Vermont", "New Hampshire", "Maine")

southeast <- c("Virginia", "Texas", "Louisiana", "Alabama", "Mississippi", "Florida", "Georgia", "South Carolina", "North Carolina")

fulldata.flw_W <- fulldata.flw %>% filter(county %in% so_cal)

```

# CA Methods

All specimen images were downloaded through the Consortium of California Herbaria (cch2.org). The search was filtered to only include specimens within *R. raphanistrum* and *R. sativus* that were collected within California and had corresponding. The full specimen data set, prior to specimen exclusions, consisted of 1,251 specimens (\***this number may be wrong and not include SD**\*.)

```{r}
library(tidyverse)
```

## Scanning Images in UCD Herbarium

I do not have notes for this :( -insert later

\~70 (**get exact number?!?!?!?**) specimens held in possession by the San Diego Natural History Museum were unimaged, but included in the full data set. These specimens were imaged using equipment (**more specificity here?**) from the University of California, Davis Center for Plant Diversity (Herbarium) and edited in Adobe Lightroom.

## Specimen Exclusions

Many of the exclusions applied to the data were guided by the exclusion criteria described in the methods of Love and Mazer (2021). 1 specimen was excluded due to the blurriness of the image. 12 images were excluded for having a poor date range recorded for the time of collection. A date range was considered unusable if it spanned more than a month due to the consistently changing *Raphanus* reproductive stages. Any specimens that were collected off of islands (e.g. Channel Islands) were excluded due to the unique climatic setting and lack of gene flow. 45 island specimens were removed from the data set based on this criterion. 2 specimens were deemed unfit for imaging, as the specimens had an envelope containing plant tissue on the specimen sheet, but no actual sample that could provide counts of reproductive structures. 71 specimens were excluded for having unreliable location data that was either difficult to assign coordinates to (see Assigning GPS Coordinates and Uncertainty) or too broad to draw accurate and precise climate data from. 5 specimens were excluded for having no listed collection date. 21 specimens were excluded on the basis of having poor metadata, lacking several pieces of information that are necessary to understand the spatial, temporal, and climatic context of the specimen. Lastly, 246 specimens were removed for possessing a spatiotemporal duplicate in the remaining data set, removing any repetition for a more cohesive data set.

## Assigning GPS Coordinates and Uncertainty

Any specimens without previously entered GPS coordinates were assigned GPS coordinates by hand. The locality data provided was used to narrow down a precise geographic position in Google Maps, occasionally with the assistance of ChatGPT or local records and articles to provide historical context. The coordinates were then taken from Google Maps and entered into a program that allows for visualizing uncertainty radii around a specified point. The uncertainty radius was hand-picked and cautiously selected to encompass the extent to which the locality description would logically still apply, similar to the approach described in the georeferencing methodology in Bontrager et. al 2025. The coordinate system for all hand-determined GPS coordinates was marked as World Geodetic System 1984 (WGS84), as used by both Google Maps and the map radius software (calcmaps.com).

WGS84 was selected as the coordinate system to standardize samples due to its prevalence in location data following its introduction in 1987. Based on information gathered from NOAA, the one to two meter difference between coordinates in NAD83 and WGS84, the two coordinate systems are considered equal for the purposes of the Raphanus data. However, the differences between NAD27 and WGS84 are often greater, so any NAD27 points were converted to WGS84. The calculations were primarily conducted en masse by ChatGPT while spot checking using the calculator provided by the West Virginia Department of Environmental Protection. The variance between coordinates in NAD27 and WGS84 was on the scale of 10^-3^ or 10^-4^, correlating to a 10-100m discrepancy between the points.

Specimens with coordinates that were listed as NAD27 in the CCH2 data were converted to CCH2. There were two subsequent pathways for specimens with coordinates that were listed as "preprocessed" or "unknown", or were otherwise unlisted. For specimens collected prior to 1987, NAD27 was broadly assumed to be the coordinate system. For specimens collected after WGS84, it was assumed that any modern GPS devices, computers, or phones operate in WGS84. Due to the uncertainty associated with these assumptions, all original coordinate and geodetic datum data was saved separately.

An additional protocol was developed for examining and assigning missing uncertainty radii entries in the CCH2 data, all of which included coordinates that served as the starting judgement point. Based on the Love and Mazer (2021) protocol, 4000 meters was deemed the ceiling cutoff for assigning uncertainty radii at this stage. For specimens that had narrowly defined, "good" locality data, the specificity of location and inclusion of coordinates generated the assumption that the uncertainty was greater than 4000 meters and the specimen's uncertainty was labeled as 3999. If a specimen was determined to have broad, "bad" location data, the specimen label was then read to determine the origin of the coordinates. If the coordinates were present on the original label, it was assumed that the coordinates are accurate and that the location data was poorly recorded, earning the specimen an uncertainty radius assignment of 3999. However, if there were no coordinates present on the label, it was assumed that they were added at a later time based on the poor location and it cannot be safely assumed that the uncertainty radius is less than 4000 meters; these specimens were then assigned an uncertainty radius of 4001. The binomial nature of this distribution can eventually be reanalyzed and assigned a more specific uncertainty value depending on the nature of the climate data that is used.

By using the 111.1 kilometer \~ 1 degree conversion, it was decided that any specimen with an original latitudinal or longitudinal coordinate that was rounded to the nearest 0.1 degree (\~11 km) should be reexamined to determine whether the coordinates or uncertainty radii could be adjusted to more precisely fall on either side of the previously set 4000 meter uncertainty radius cutoff. Specimens with location descriptions that were determined to be broader in extent than 4000 meters, were assigned an uncertainty radius of 4001 if their previously listed uncertainty radius was not already greater than 4000. Specimens with location data that was deemed more specific than 4000 meters (e.g. parks, specific intersections, street addresses) then received adjusted, more precise coordinates in WGS84 and were reassigned a fitting uncertainty radius using the map radius software found at calcmaps.com.

## Choosing the Right Machine Learning Program

Many previously published scientific papers have also utilized images from various herbaria or citizen science platforms such as iNaturalist to study phenological trends. These papers were examined to get a sense of the software options available to annotate images, build models, and subsequently train machine learning programs to apply the model to a larger data set. There were many possible paths that varied in coding language, accessibility, complexity, and suitability for the needs of the project. Eventually, Roboflow was selected as the most practical choice for conducting the study, as the software provides the ease of completing annotation and training models in the same location. Additionally, Roboflow provides model evaluations with specific feedback such as precision and recall metrics as well as confusion matrices to easily visualize errors in the model and coding prompts to deploy the model to other sets of images.

## Pulling Images from CCH2 for Model Evaluation

```{python}

#| eval: false
#| label: Code to Pull Images from CCH2 data file

# pip install requests beautifulsoup4

import requests
from bs4 import BeautifulSoup
import os

# Create a folder to store downloaded images
os.makedirs("cch2_images", exist_ok=True)

# Load URLs from file
with open("specimen_urls.txt") as f:
    urls = [line.strip() for line in f if line.strip()]

for url in urls:
    try:
        # Get the HTML content of the page
        response = requests.get(url)
        response.raise_for_status()

        soup = BeautifulSoup(response.text, 'html.parser')

        # Look for the image (usually inside <img> with id or class)
        img_tag = soup.find("img", {"id": "imgScan"})

        if img_tag and img_tag.get("src"):
            img_url = img_tag["src"]
            if img_url.startswith("/"):
                img_url = "https://cch2.org" + img_url

            # Get a unique filename
            occid = url.split('=')[-1]
            filename = f"cch2_images/{occid}.jpg"

            # Download the image
            img_data = requests.get(img_url).content
            with open(filename, 'wb') as f:
                f.write(img_data)

            print(f"Downloaded image for occid={occid}")

        else:
            print(f"No image found at {url}")

    except Exception as e:
        print(f"Error processing {url}: {e}")



```

## Selecting Specimens for Model Training

It was initially decided that 100 specimens, slight greater than ten percent of the full data set, would be selected to be annotated and used to train the phenology models. The first consideration made was to include at least one specimen from each of the 31 herbaria that are represented by the complete specimen data set. Afterwards, specimens were added in to be relatively proportional to the distribution of the data set. When continuing to add in specimens after initial herbarium representation was accounted for, three factors received the majority of focus: time of collection, reproductive stage, and complexity of specimens. The time of collection was critical to creating a diverse, representative training set. While most specimens are collected during the spring and summer months, when flowering and fruiting is widespread, specimens collected during both the very beginning and the later months of the calendar year were fairly represented. Additionally, a majority of specimens possess both flowering and fruiting structures, so training set selection was carefully done to also include specimens that had not yet begun to fruit or that only had fruit left. Lastly, due to the many different conditional circumstances that can occur in herbarium specimens, such as image quality, age, human error, etc., both complex, overlapping specimens and well-defined clear cut specimens were selected for annotation to prepare the model for any unusual specimen appearances.

## Annotating Herbarium Specimens

An initial 100 specimens were uploaded into Roboflow to be annotated. The Roboflow software has a file size limit of 20 MB; any specimen images that exceeded that size were shrunk using code written in terminal:

```{bash}
#| eval: false
#| label: Code to Resize Images that were too big for Robolfow

# image resizing code:

# number is pixel width, base line:
sips --resampleWidth 5000 input.jpg --out SML_output.jpg

for file in *.jpg; do
  sips --resampleWidth 5000 "$file" --out "SML_$file"
done
```

Specific procedures and approaches were developed after discussing the logistics and potential inconsistencies that could occur when training a Machine Learning model to recognize the three different reproductive classes: bud clusters, flowers, and fruit. Bud clusters fall along a spectrum of appearances. Bud clusters were annotated by outlining groups of, or individual buds, that were clearly young and appeared green and fuzzy or were slightly older and had a darker brown, woody looking appearance. By this guideline, it was acceptable to annotate just a portion of the bud cluster that met the brown, woody appearance that could be learned by the model, thereby representing the entire cluster as a whole.

To prevent model confusion between bud clusters and flowers, any buds where petals had begun to emerge and both bud cluster tissue and flower tissue were visible were excluded. Flowers were annotated individually as best as possible, even if clustered together or pressed to the sheet in a visually confusing manner. Flowers were annotated without the pedicel. Any flowers that had broken off but were still glued onto the specimen sheet were included. Flowers that seemed to have undergone senescence (shriveled, darker and drier in appearance) were excluded, as they were not contributing to the model's reproductive output at the time of collection.

Fruits were annotated with the pedicel to provide consistency to the model in identifying the structure. Immature and mature fruits vary drastically in length, width, and curvature, so providing the pedicel was intended to help the model identify fruits. Like flowers, fruits that were partially or completely broken off but included on the specimen sheet were included in annotations.

Empty pedicels were not included in annotations (somewhat due to time constraints). Pedicels could be identified as corresponding to either flowers or fruits depending on their location on the plant and proximity to developed and fruits. However, for our model annotations, empty pedicels were omitted.

Leaves were not annotated for our model (again time constraints in place), though there would be some benefits to doing so. Although leaves are not a reproductive structure, including them in annotations could potentially help the model differentiate between leaves and fruits. By teaching the model what fruits and leaves are separately, it could have more success in differentiating the two when run.

## Training Model Versions

**Insert text here about how we picked different models to train, how each version did, the 70-20-10 distribution, model metrics and evaluations, etc.**

## Deploying Model from Roboflow

```{bash}
#Code for doing Roboflow model to JSONS 

import json
import os
from concurrent.futures import ThreadPoolExecutor, as_completed
from pathlib import Path

from inference_sdk import InferenceHTTPClient
from PIL import Image

# Do not hardcode keys in this file; export before running:
#   export ROBOFLOW_API_KEY="your_key"
API_KEY = os.environ.get("ROBOFLOW_API_KEY")
if not API_KEY:
    raise SystemExit(
        "Set ROBOFLOW_API_KEY in the environment, e.g. export ROBOFLOW_API_KEY=..."
    )

SCRIPT_DIR = Path(__file__).resolve().parent
MODEL_ID = "phenologyscoringeve/10"

CLIENT = InferenceHTTPClient(
    api_url="https://serverless.roboflow.com",
    api_key=API_KEY,
)


def process_image(image_path: Path):
    """Run inference and save output to JSON.

    Pillow images are used on purpose: with serverless API v0 the SDK otherwise
    uses OpenCV (imread/imencode), which is not thread-safe across workers and
    can yield corrupted base64 under ThreadPoolExecutor.
    """
    with Image.open(image_path) as im:
        rgb = im.convert("RGB")
        result = CLIENT.infer(inference_input=rgb, model_id=MODEL_ID)
    out_file = image_path.with_suffix(".json")
    with open(out_file, "w") as f:
        json.dump(result, f, indent=4)
    return image_path.name, out_file.name


def main():
    files = sorted(SCRIPT_DIR.glob("*.jpg"))
    if not files:
        print(f"No .jpg files in {SCRIPT_DIR}")
        return

    max_workers = min(5, len(files))
    with ThreadPoolExecutor(max_workers=max_workers) as executor:
        futures = {executor.submit(process_image, p): p for p in files}
        for future in as_completed(futures):
            path = futures[future]
            try:
                in_name, out_name = future.result()
                print(f"Processed {in_name} -> {out_name}")
            except Exception as exc:
                print(f"FAILED {path.name}: {exc}")


if __name__ == "__main__":
    main()


```

```{python}
#Code for converting model output JSON to Dataframe
import glob
import json
import pandas as pd
from collections import Counter

# Desired class order
class_order = ["Bud Cluster", "Flower", "Fruit"]

rows = []

# Loop through all JSON files
for file in glob.glob("*.json"):
    with open(file, "r") as f:
        result = json.load(f)
    
    # Extract class names from predictions
    classes = [pred["class"] for pred in result.get("predictions", [])]
    
    # Count how many of each class were detected
    counts = Counter(classes)
    
    # Image ID without extension
    image_id = file.replace(".json", "")
    
    # Build row dict with all classes in fixed order
    row = {"image": image_id}
    for cls in class_order:
        row[cls] = counts.get(cls, 0)
    
    rows.append(row)

# Build DataFrame with class order
df = pd.DataFrame(rows, columns=["image"] + class_order)

# Set image as index if desired (optional)
# df = df.set_index("image")

# Save to CSV
df.to_csv("missing_inference_counts.csv", index=False)

print("Saved dataset to missing_inference_counts.csv")
print(df)
```

## Converting Model Data into a Usable Format

The following code was created to achieve two purposes. The first was to combine the JSON information on phenology (reproductive structure counts) after running both the original 155 annotated specimen images and the remaining 782 unannotated specimen images through the Roboflow model. The second was to combine all the phenology data with a subset of the data provided by CCH2. Intermediate steps were incorporated as needed to ensure compatibility when merging data frames.

```{r}
#| label: Combining CCH2 Data with Roboflow Scores

#Load tidyverse package into R from Library
library(tidyverse)

#Simplify CCH2 data down to the columns of interest
#Columns of interest were determined based on what would be necessary to identify each specimen and what would be necessary for using climate data to understand the effects of climate variables on flowering time
important_specimen_data <- read.csv("Input_Files/FullSpecimenCCH2Data.csv") %>% select(id, 
institutionCode, specificEpithet, recordedBy, eventDate, year:day, 
verbatimEventDate, county, locality, 
decimalLatitude:coordinateUncertaintyInMeters)

#Change "id" column from integer class to character class in CCH2 data frame
important_specimen_data$id <- as.character(important_specimen_data$id)

#Combine Annotated and Unannotated phenology class data into one big data frame
ann <- read.csv("Input_Files/annotated_inference_counts.csv")
unann <- read.csv("Input_Files/inference_counts_RETRY.csv")
entire_class_dataset <- bind_rows(ann, unann)

#Change the name of "image" column in phenology class data frame to "id" to align with CCH2 data frame
entire_class_dataset <- entire_class_dataset %>% rename(id = image)

#Images that were larger than Roboflow's maximum upload size of 20MB were resized via code and given the prefix "SML_" in the filename to discern the edited image. However, this prefix appeared in the id column of the phenology class data and needed to be removed in order to combine the phenology class and CCH2 data frames.
entire_class_dataset$id <- str_replace(entire_class_dataset$id, "^SML_", "")

#Remove any observation (e.g. duplicates) that is in phenology class, but not CCH2 data
rm <- setdiff(entire_class_dataset$id, important_specimen_data$id)
entire_class_dataset <- entire_class_dataset %>% filter(!entire_class_dataset$id 
                                                        %in% rm)

#Merge the CCH2 data frame and the full phenology class data frame into one large data frame, matching based on the id column with a precautionary full join
totalinfo <- full_join(entire_class_dataset, important_specimen_data, by = "id")

#Export the full combined data frame into a CSV for further uses
write.csv(totalinfo, "Output_Files/completed_specimen_data.csv", row.names = FALSE)

```

At a later time, it was discovered that, when running the code for the model on specimens, there was one previously hand-annotated specimen that was given counts of 0 bud clusters, 0 flowers, and 0 fruits. After hand checking this image, it was clear that all three reproductive structures could be found on the specimen. The original hand-annotation counts from Roboflow were assigned to the specimen using the following code:

```{r}
#| eval: false
# #Read in full complete data CSV
# fulldata<- read.csv("Output_Files/completed_specimen_data.csv")
# 
# #Fix the specimen that was falsely determined to have no assigned reproductive structure counts by individually assigning the accurate counts from the original annotations done by hand in Roboflow
# fulldata <- fulldata %>%
#      mutate(
#            Bud.Cluster = if_else(id == 1368028, 3, Bud.Cluster),
#         Flower = if_else(id == 1368028, 2, Flower),
#          Fruit = if_else(id == 1368028, 2, Fruit))
# 
# #Export this adjusted change back into the CSV
# write.csv(fulldata, "Output_Files/completed_specimen_data.csv", row.names = FALSE)
```

## Extracting Climate Data

**Add in text here about how climate NA was Chosen and how we sourced data, lots of credit to Bontrager, etc.**

```{r}
#EVENTUAL BONTRAGER + Additional to us code here... unless we somehow link the entire other markdown into here if that's a thing
```

## Plotting and Figure Creation

#### Frequency of Occurrence for all Reproductive Structures

```{r}
#| eval: false

#Load in Dplyr and GGplot2
library(dplyr)
library(ggplot2)

#Read in full complete data CSV, and add column with month abbreviations in text
fulldata<- read.csv("Output_Files/completed_specimen_data.csv") %>% mutate(monthname = month.abb[as.numeric(month)])


#Create frequency histogram for Bud Clusters
hist(fulldata$Bud.Cluster, breaks = 10, freq = TRUE, col = "darkorchid3", 
     border = "white", main = "Bud Cluster Frequency in California Raphanus 
     Herbarium Specimens", ylab = "Frequency", xlab = "# of Bud Clusters")
```

The majority of specimens had just one bud cluster, with a steep drop off between one and two bud clusters. This is likely caused by both model inaccuracies and the viusally ambiguous appearance of bud clusters. In all, the current model cannot reliably produce individual counts for bud clusters on *Raphanus* specimens.

```{r}
#| eval: false

#Create frequency histogram for Flowers
hist(fulldata$Flower, breaks = 40, freq = TRUE, col = "firebrick3", 
     border = "white", main = "Flower Frequency in California Raphanus 
     Herbarium Specimens", ylab = "Frequency", xlab = "# of Flowers")
```

There is a right skewed distribution for the flower frequency curve, with most observations falling between 0-10 flowers. This data could further to be studied how flower number and flowering time varies across latitudes in California, as different latitudes experience vastly different climates. The presence or absence of flowers is a much more reliable metric that can be gauged from the model. While the individual counts for flowers are less accurate, specimens can answer the binomial question: "Does the specimen have flowers," providing insight into broader phenological stage.

```{r}
#| eval: false

#Create frequency histogram for Fruits
hist(fulldata$Fruit, breaks = 25, freq = TRUE, col = "seagreen1", 
     border = "black", main = "Fruit Frequency in California Raphanus 
     Herbarium Specimens", ylab = "Frequency", xlab = "# of Fruits")
```

Similar to flowers, the distribution curve for fruits is skewed right, with over 25% of the observations having between 0-5 fruits. Similar analyses can be conducted for fruits based on what is done with flowers to get a sense of broader phenological stage and variances in latitude, but this is not necessary.

#### Frequency of Flowering Specimens in Each Month

```{r}
#| label: Creating Histogram of Flowering Specimens per month


#Subset full data set to only include flowering specimens
flowering <- fulldata %>% filter(Flower > 0)


#Create a Histogram for flowering specimens with month on the X-Axis and Frequency(# of specimens) on the Y-axis

full_histo <- ggplot(flowering, aes(x = month)) + 
  #Set bin width and change the color of the histogram
  geom_histogram(binwidth = 1, fill = "lightsteelblue2", color = "black") + 
  #Label X-axis with month abbreviations
  scale_x_continuous(breaks = 1:12, labels = month.abb) + 
  #Add axis and graph titles
  labs(x = "Month", y = "Frequency", title = "Frequency of flowering specimens \ncollected in each month ") + 
  #Make the background blank with no grid lines and center the grpah title
  theme(plot.title = element_text(hjust = 0.5), panel.grid.minor = element_blank(), panel.border = element_rect(fill = "white"), panel.background = element_rect(fill = "white")) +
  #Add a visible X-Axis
  geom_hline(yintercept = 0, color = "black", linewidth = 0.75)

#Save the Figure as a PDF
#ggsave("Figures/Full_Histo_2.pdf", device = pdf, plot = full_histo, height = 4, width = 5, units = "in", dpi = 400)
```

By looking at this graph, we can tell that the majority of specimens that were flowering were collected in the Spring months between March and May. Over 250 flowering specimens were collected in April alone. When reading through the scientific articles, I was intrigued by the Ridley and Ellstrand 2009 papers, where they had 2 study sites for both southern inland and northern coastal specimens to be grown and studied. One specimen site was inland and southern, at UCR, and the other was coastal and northern, at Point Reyes. The authors mentioned that at the Point Reyes site, radishes tend to behave as a summer annual, flowering near the end of summer, while at the UCR site the Radish behave like winter annuals and flower in early spring. I think it could be interesting to expand on this histogram and incorporate location data to try and further understand this aforementioned trend and see if there is proof of this idea in the herbarium specimens.

```{r}
#| eval: false


#Filter for specimens that are fruiting
fruiting <- fulldata %>% filter(Fruit > 0)

#Create a Histogram for fruiting specimens with month on the X-Axis and Frequency(# of specimens) on the Y-axis

full_histo_fruit <- ggplot(fruiting, aes(x = month)) + 
  #Set bin width and change the color of the histogram
  geom_histogram(binwidth = 1, fill = "lightsteelblue2", color = "black") + 
  #Label X-axis with month abbreviations
  scale_x_continuous(breaks = 1:12, labels = month.abb) + 
  #Add axis and graph titles
  labs(x = "Month", y = "Frequency", title = "Frequency of fruiting specimens \ncollected in each month ") + 
  #Make the background blank with no grid lines and center the grpah title
  theme(plot.title = element_text(hjust = 0.5), panel.grid.minor = element_blank(), panel.border = element_rect(fill = "white"), panel.background = element_rect(fill = "white")) +
  #Add a visible X-Axis
  geom_hline(yintercept = 0, color = "black", linewidth = 0.75)

#Save the Figure as a PDF
#ggsave("Figures/Full_Histo_Fruit.pdf", device = pdf, plot = full_histo_fruit, height = 4, width = 5, units = "in", dpi = 400)



```

A similar trend follows when looking at the graphs of fruit-month frequency and flower-month frequency with most specimens that were fruiting being collected in the spring, specifically April. Fruiting time is not as prevalent in literature as flowering time.

#### Temporal Extremity Flowering Plots:

```{r}
#| label: Creating Histogram of Flowering Specimens per month pre-1950 and post-2000


#Load in relevent packages
library(dplyr)
library(ggplot2)
library(patchwork)
library(lubridate)
library(colorblindcheck)


# Subset the full dataset to be just flowering specimens, add month abbreviation column
fulldata.flw <- read.csv("Output_Files/completed_specimen_data.csv") %>% mutate(monthname = month.abb[as.numeric(month)]) %>% filter(Flower > 0)

#Check to see approximate number of specimens per year via Histogram
#hist(fulldata.flw$year, breaks=60)

# Clean Up County Names for Uniformity
 fulldata.flw$county <- gsub("\\s*[Cc]ounty\\s*$", "", fulldata.flw$county) 
# remove the phrase "county"
fulldata.flw$county <- trimws(fulldata.flw$county) # trim whitespace

#Filter for the selected Southern California Counties
so_cal <- c("Santa Barbara", "San Luis Obispo", "San Diego", "Los Angeles", "Orange", "Riverside", "San Bernardino", "Ventura")

fulldata.flw_so_cal <- fulldata.flw %>% filter(county %in% so_cal)


#Separate the data into two separate time periods (pre-1950 ="old" and post-2000 = "recent")

old_so_cal <- fulldata.flw_so_cal %>% filter(year < 1950 ) %>% mutate(day = replace(day, day == 0, 1))
old_so_cal$date <- make_date(old_so_cal$year, old_so_cal$month, old_so_cal$day)
old_so_cal$DOY <- yday(old_so_cal$date)

#mean(old_so_cal$DOY)
#tmp_o <- old_so_cal %>% filter(DOY < 213)
#mean(tmp_o$DOY)
#hist(old_so_cal$DOY, breaks =40)
#abline(v=mean(old_so_cal$DOY), col='red')

recent_so_cal <- fulldata.flw_so_cal %>% filter(year > 2000) %>% mutate(day = replace(day, day == 0, 1))
recent_so_cal$date <- make_date(recent_so_cal$year,recent_so_cal$month, recent_so_cal$day)
recent_so_cal$DOY <- yday(recent_so_cal$date)

#mean(recent_so_cal$DOY)
#tmp <- recent_so_cal %>% filter(DOY < 213)
#mean(tmp$DOY)
#hist(recent_so_cal$DOY, breaks =40)
#abline(v=mean(recent_so_cal$DOY), col='red')


#Colorblind Check for Histogram Colors
#colorblindcheck::palette_plot(c("black", "#825E89", "lightsteelblue2"))


#Create a Histogram where X-Axis is month and Y-axis is frequency/number of specimens pre-1950
pre_histo <- ggplot(data = old_so_cal, aes(x = factor(month, levels = 1:12) )) + #Make a bar graph for discrete variables and choose graph color
  geom_bar(color = "black", fill = "#825E89", width = 1) + 
  #Make the X-axis labels appear as month abbreviations
  scale_x_discrete(labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"), drop = FALSE) +
  # Add axis and graph titles
  labs(x = "Month", y = "Frequency", title = "Frequency of flowering specimens \ncollected in southern California pre-1950") + # Center the title in the middle, make background blank wihtout grid lines
  theme( plot.title = element_text(hjust = 0.5), panel.grid.minor = element_blank(), panel.border = element_rect(fill = "white"), panel.background = element_rect(fill = "white")) +
  #Add visible X-axis
  geom_hline(yintercept = 0, color = "black", linewidth = 0.75) #+
  
  #Add mean bar for day of year collected
  #geom_vline(xintercept = )

#Save the Histogram as a pdf
#ggsave("Figures/Pre-1950-Histo.pdf", device = pdf, plot = pre_histo, height = 4, width = 5, units = "in")

#Create a Histogram where X-Axis is month and Y-axis is frequency/number of specimens post-2000
post_histo <- ggplot(data = recent_so_cal, aes(x = factor(month, levels = 1:12) )) + #Make a bar graph for discrete variable and choose graph color
  geom_bar(color = "black", fill = "#825E89", width = 1) + 
  #Make the X-axis labels appear as month abbreviations
  scale_x_discrete(labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"), drop = FALSE) +
  # Add axis and graph titles
  labs(x = "Month", y = "Frequency", title = "Frequency of flowering specimens \ncollected in southern California post-2000") + 
  #Center the title in the middle, make background blank without gridlines
  theme(plot.title = element_text(hjust = 0.5), panel.grid.minor = element_blank(), panel.border = element_rect(fill = "white"), panel.background = element_rect(fill = "white")) +
  #Add visible X-axis
  geom_hline(yintercept = 0, color = "black", linewidth = 0.75)

#Save the histogram as a PDF
#ggsave("Figures/Post-2000-Histo.pdf", device = pdf, plot = post_histo, height = 4, width = 5, units = "in")

```

#### Temporal Extremity Specimen Maps

```{r}
#| label: Creating Maps of Flowering Specimens pre-1950 and post-2000


#Load the maps package
library(maps)

# Get a California Map
state <- map_data("state")
california <- subset(state, region == "california")


#Subset the data for only the Southern California counties selected and further subset into pre-1950 and post-2000 collected specimens

old <- fulldata.flw %>% filter(year < 1950 ) %>% filter(Flower > 0) %>% mutate( highlight = ifelse(county %in% so_cal, "Southern California", "Not Included")) 

recent <- fulldata.flw %>% filter(year > 2000) %>% filter(Flower > 0) %>% mutate( highlight = ifelse(county %in% so_cal, "Southern California", "Not Included"))

#Create a map showing specimens collected in So Cal pre-1950

pre2 <- ggplot() + 
  #Create California Outline
  geom_polygon(data = california, aes(x = long, y = lat, group = group), fill = "lightsteelblue2", color = "white") + 
  #Add in Old Specimen Data and color based on highlight column
  geom_point(data = old, aes(x = decimalLongitude, y = decimalLatitude, color = highlight)) + 
  #Label the Axes
  labs( x = "", y = "") + theme(axis.ticks = element_blank(), axis.text = element_blank()) + scale_color_manual(values = c("Southern California" = "#825E89", "Not Included" = "black"), name = "Specimens")

#Save the Map as a pdf
#ggsave("Figures/pre_1950_map_2.pdf", device = "pdf" , plot = pre2, height = 2, width = 4, units = "in")

#Create a map showing specimens collected in So Cal post-2000

post2 <- ggplot() + 
  #Create California Outline
  geom_polygon(data = california, aes(x = long, y = lat, group = group), fill = "lightsteelblue2", color = "white") + 
  #Add in Recent Specimen Data and color based on highlight column
  geom_point(data = recent, aes(x = decimalLongitude, y = decimalLatitude, color = highlight)) + 
  #Label the Axes
  labs( x = "", y = "")  + theme(axis.ticks = element_blank(), axis.text = element_blank()) + scale_color_manual(values = c("Southern California" = "#825E89", "Not Included" = "black"), name = "Specimens")


#Save the map as a pdf
#ggsave("Figures/post_2000_map_2.pdf", device = "pdf" , plot = post2, height = 2, width = 4, units = "in")
```
